---
title: "45-lmm1-reproducibility"
output: html_notebook
---

```{r setup, echo=FALSE, message=FALSE}
bookdown_compile <- FALSE

library(ggplot2)
library(ggpubr)
library(cowplot)
library(readxl)
library(emmeans)
library(data.table)
library(lmerTest)
library(harrellplot)
library(car)
library(nlme)

if(bookdown_compile==TRUE){
  data_path <- "data" # bookdown
  source("R/clean_label.R") # bookdown
}else{
  data_path <- "../data" # notebook
  source("../R/clean_label.R")
}

```
# parasite
```{r parasite-import}
folder <- "Data from Synergistic interaction between UVB radiation and temperature increases susceptibility to parasitic infection in a fish"
fn <- "Cramp et al raw data.xlsx"
sheet_i <- "Infection Intensity"
file_path <- paste(data_path, folder, fn, sep="/")
parasite <- data.table(read_excel(file_path, sheet=sheet_i))
colnames(parasite) <- clean_label(colnames(parasite))
setnames(parasite, old="Whitespots/fish", new="Whitespots")
parasite[, Tank:=factor(Tank)]
n_fish_per_tank <- parasite[, .(N=.N), by=.(UV_Level, Temperature, Tank)]
n_tanks_per_combo <- n_fish_per_tank[, .(N=.N), by=.(UV_Level, Temperature)]
```

```{r parasite-analysis}
# what authors did
fit <- lmer(Whitespots ~ UV_Level*Temperature + (1|Tank), data=parasite)
parasite[, Whitespots_res1:=residuals(fit)]
parasite[, Whitespots_fitted:=fitted(fit)]
ggqqplot(data=parasite, x = "Whitespots_res1", title="Normal Q-Q")
qplot(data=parasite, x=Whitespots_fitted, y=sqrt(abs(Whitespots_res1))) + 
  geom_smooth(method="lm")

anova(fit)
coef(summary(fit))

# poisson

# negbinom
  fit.full <- glmer.nb(Whitespots ~ UV_Level*Temperature + (1|Tank), data=parasite)
  simulationOutput <- simulateResiduals(fittedModel=fit.full, n=n_sim, refit=TRUE)
  plot(simulationOutput, asFactor = F)
  testDispersion(simulationOutput)
  
fit.red <- glmer.nb(Whitespots ~ UV_Level+Temperature + (1|Tank), data=parasite)
anova(fit.full, fit.red)
coef(summary(fit))

```

```{r parasite-pooling}
complete_pooling <- lm(Whitespots ~ UV_Level*Temperature, data=parasite)
no_pooling <- lmList(Whitespots ~ UV_Level*Temperature | Tank, data=parasite)
partial_pooling <- lmer(Whitespots ~ UV_Level*Temperature + (1|Tank), data=parasite)
aggregate_pooling <- lm(Whitespots ~ UV_Level*Temperature, data=parasite[, .(Whitespots=mean(Whitespots)), by=.(UV_Level, Temperature, Tank)])

```

# coral ("When environmental factors become stressors...")

```{r coral-read}
folder <- "Data from When environmental factors become stressors- interactive effects of vermetid gastropods and sedimentation on corals"
fn <- "VermetidSedimentData_ZillGilOsenberg_DRYAD.xlsx"
sheet_i <- "Coral Growth Rate Data"
file_path <- paste(data_path, folder, fn, sep="/")
coral <- data.table(read_excel(file_path, sheet=sheet_i))
setnames(coral, old=colnames(coral), new=clean_label(colnames(coral)))
coral[, Vermetids:=factor(Vermetids)]
coral[, Sediment:=factor(Sediment)]
```

```{r coral-lmm}
# identify outlier
fit.lm <- lm(GrowthRate ~ Vermetids*Sediment, data=coral)
# plot(fit.lm)
# observation 2 looks like the outlier and its in the +Sedimentation treatment

inc <- c(1, 3:nrow(coral))

fit.lme <- lme(GrowthRate ~ Vermetids*Sediment, random= ~1|Block, data=coral[inc])
anova(fit.lme, type="marginal")

fit.lmer <- lmer(GrowthRate ~ Vermetids*Sediment + (1|Block), data=coral[inc])
anova(fit.lmer)


```

# Mantid ("Impacts of silicon-based grass defence...")

Data from Impacts of silicon-based grass defences across trophic levels under both current and future atmospheric CO2 scenarios

```{r mantid-read}
folder <- "Data from Impacts of silicon-based grass defences across trophic levels under both current and future atmospheric CO2 scenarios"
fn <- "Supplementary material data.xlsx"
sheet_i <- "Mantid performance"
file_path <- paste(data_path, folder, fn, sep="/")
mantid <- data.table(read_excel(file_path, sheet=sheet_i))
setnames(mantid, old=colnames(mantid), new=clean_label(colnames(mantid)))

# relabel factors
mantid[, CO2_cont:=CO2_concentration]
mantid[, CO2:=factor(CO2_concentration, c("400", "640", "800"))]
mantid[, Si:=factor(Silicon_treatment)]

# transformations
```

```{r mantid}
# one observation of $mass_of_crickets_eaten is negative, which makes me think the archived data are log transformed. Stating the transformation is "exp" is confusion

mantid[, .(mass_of_crickets_eaten=mean(exp(mass_of_crickets_eaten))), by=.(CO2, Si)]
mantid[, .(mass_of_crickets_eaten=mean((mass_of_crickets_eaten))), by=.(CO2, Si)]
# Fig 2 would have to be on log scale

# mass of crickets eaten
# are the data already log transformed?
# df matches when CO2 is treated as factor (note treated as continuous in Fig. 2)
inc <- which(mantid$mass_of_crickets_eaten > 0)
fit_lme <- lme(mass_of_crickets_eaten ~ CO2*Si, random= ~1|Mantid_ID, data=mantid)
anova(fit_lme)
anova(fit_lme, type="marginal")
Anova(fit_lme, type="3")
fit_lme <- lme(exp(mass_of_crickets_eaten) ~ CO2*Si, random= ~1|Mantid_ID, data=mantid)
anova(fit_lme)
anova(fit_lme, type="marginal")
Anova(fit_lme, type="3")
fit_lme <- lme(log(mass_of_crickets_eaten) ~ CO2*Si, random= ~1|Mantid_ID, data=mantid[inc])
anova(fit_lme)
fit_lme <- lme(mass_of_crickets_eaten ~ CO2_cont*Si, random= ~1|Mantid_ID, data=mantid)
anova(fit_lme)
fit_lme <- lme(exp(mass_of_crickets_eaten) ~ CO2_cont*Si, random= ~1|Mantid_ID, data=mantid)
anova(fit_lme)
fit_lme <- lme(log(mass_of_crickets_eaten) ~ CO2_cont*Si, random= ~1|Mantid_ID, data=mantid[inc])
anova(fit_lme)

# frass
mantid[, .(frass_mass=mean(frass_mass*1000),
           frass_mass_sq=mean(frass_mass^2)),
       by=.(CO2, Si)]
# Fig 2 would have to be on log scale

# already sqrt transformed?
fit_lme <- lme(frass_mass ~ CO2*Si, random= ~1|Mantid_ID, data=mantid)
anova(fit_lme)
emmeans(fit_lme, specs=c("CO2", "Si"))

fit_lme <- lme(sqrt(frass_mass) ~ CO2*Si, random= ~1|Mantid_ID, data=mantid)
anova(fit_lme)

data.table(pre=(mantid$pre_cricket_mass), check=(mantid$mass_of_crickets_eaten + mantid$cricket_remains + mantid$frass_mass))
data.table(pre=exp(mantid$pre_cricket_mass), check=exp(mantid$mass_of_crickets_eaten) + exp(mantid$cricket_remains) + exp(mantid$frass_mass))
data.table(pre=exp(mantid$pre_cricket_mass), check=exp(mantid$mass_of_crickets_eaten) + exp(mantid$cricket_remains) + (mantid$frass_mass)^2)

```

# Heating ("Above and below ground responses of four tundra plant functional types...")

Data from Above and below ground responses of four tundra plant functional types to deep soil heating and surface soil fertilization

```{r lmm1-AG_biomass, echo=FALSE, eval=FALSE}
folder <- "Data from Data from Above and below ground responses of four tundra plant functional types to deep soil heating and surface soil fertilization"
fn <- "2015.05.18_weighing results aboveground.xlsx"
file_path <- paste(data_path, folder, fn, sep="/")
AG_biomass <- data.table(read_excel(file_path, range="A1:F352"))
AG_biomass[, log_weight:=log(weight)]
AG_biomass[, log_weight_p1:=log(weight+1)]

# make block and pft factors
AG_biomass[, block:=factor(block)]
AG_biomass[, pft:=factor(pft)]

# change plot to unique labels so that labels differ among blocks
AG_biomass[, plot2 :=factor(paste(block,plot,sep='-'))]

# heating levels
# C - "control", PC - "cable", or procedural control, H=warming
heating_levels <- c("H-", "PC", "H+")
# fertlization levels
# C - "control"
# F - "fertilized"
fertlization_levels <- c("F-", "F+")
AG_biomass[, Heating:=ifelse(plot==3 | plot==4, heating_levels[1],
               ifelse(plot==2 | plot==6, heating_levels[2],
               ifelse(plot==1 | plot==5, heating_levels[3], NA))
                      )]
AG_biomass[, Heating:=factor(Heating, heating_levels)]
AG_biomass[, Fertilization:=ifelse(plot %in% c(2, 4, 5), fertlization_levels[1], fertlization_levels[2])]
AG_biomass[, Fertilization:=factor(Fertilization, fertlization_levels)]

# combine C + PC
AG_biomass[, Heating2:=ifelse(as.character(Heating)=="PC", "C", as.character(Heating))]
AG_biomass[, Heating2:=factor(Heating2)]
# plot
# 1: warming & fertilization
# 2: cable
# 3: fertilization
# 4: control
# 5: warming
# 6: cable & fertilization
# 
# PFT
# DS: deciduous shrub (Betula nana, Salix glauca, S. pulchra)
# ES: evergreen shrub (Ledum decumbens, Vaccinium vitis-idaea)
# Gr: grass (Arctagrostis latifolia, A. calamagrostis)
# Se: sedge (Eriophorum vaginatum, Carex bigelowi)
```

```{r lmm1-AG-biomass-plot, echo=FALSE, eval=FALSE}
hp_grass <- harrellplot(x="Heating", 
                        y="log_weight", 
                        g="Fertilization", 
                        rintcols="block",
                        fit.model = "lmm",
                        data=AG_biomass[Heating!="PC" & pft=="gr"],
                        contrasts.method="revpairwise", 
                        rel_height = 1
                        )
hp_grass$gg
```

```{r lmm1-AG-biomass-model, echo=FALSE, eval=FALSE}
# Heating df is 1, but there are three levels so they have either dropped "Cable" or combined it with control

#plot as coded, combining cable + control
fit1 <- lmer(log_weight_p1 ~ Heating2*Fertilization*pft + (1|block/plot), data=AG_biomass)
anova(fit1)

fit2 <- lmer(log_weight_p1 ~ Heating2*Fertilization*pft + (1|block), data=AG_biomass)
anova(fit2)

fit3 <- lmer(log_weight_p1 ~ Heating2*Fertilization*pft + (1|block/plot/point), data=AG_biomass)
anova(fit3)

#plot as coded, dropping cable
fit4 <- lmer(log_weight_p1 ~ Heating*Fertilization*pft + (1|block/plot),
             data=AG_biomass[Heating!="PC"])
anova(fit4)
fit5 <- lmer(log_weight ~ Heating*Fertilization*pft + (1|block/plot),
             data=AG_biomass[Heating!="PC"])
anova(fit5)
fit6 <- lmer(log_weight ~ Heating2*Fertilization*pft + (1|block/plot),
             data=AG_biomass)
anova(fit6)
fit7 <- lmer(log_weight_p1 ~ Heating*Fertilization*pft + (Heating|block/plot) + 
               (Fertilization|block/plot), data=AG_biomass[Heating!="PC"])
anova(fit7)






```

```{r lmm1-AG-biomass-grass, eval=FALSE}
fit1 <- lmer(log_weight_p1 ~ Heating2*Fertilization + (1|block),
             data=AG_biomass[pft=="gr"])
anova(fit1)
fit2 <- lmer(log_weight_p1 ~ Heating*Fertilization + (1|block),
             data=AG_biomass[pft=="gr" & Heating!="PC"])
anova(fit2)

```
```
