---
title: "nested models"
output: html_notebook
---

```{r}
library(lme4)
data(Pastes)
```

1. 10 batches with 3 casks per batch and 2 tests per cask.

2. batch with 10 levels

3. cask with 3 levels. These have the same name within each batch ("a", "b", "c") but "a" is not meaningful, hence nested not crossed

4. sample with 30 levels reflecting all combinations of batch x cask. Hence this is an interaction batch:cask. If we gave the casks different names across batches this would be the same as cask.

All of these specifications are the same.
```{r}
# classic "nested" specification
m1 <- lmer(strength ~ 1 + (1|batch/cask), data=Pastes)
# lmer expands the nested specification to one with the interaction. Then these are simply additive random effects
m2 <- lmer(strength ~ 1 + (1|batch) + (1|batch:cask), data=Pastes)
# "hidden" interaction unless one realizes that sample is all combinations of batch*cask
m3 <- lmer(strength ~ 1 + (1|batch) + (1|sample), data=Pastes)
# this would be specification in m2 but if we labeled casks differently among batches
m4 <- lmer(strength ~ 1 + (1|batch) + (1|batch:sample), data=Pastes)
# and the same as m1 but if we labeled casks differently among batches.
m5 <- lmer(strength ~ 1 + (1|batch/sample), data=Pastes)
summary(m1)
summary(m2)
summary(m3)
summary(m4)
summary(m5)
```

