# Analyses for Figure 2 of "ASK1 inhibits browning of white adipose tissue in obesity" {-}

## Setup

Some plots in this document require the file "ggplotsci.R" within the "R" folder, within the project folder.

```{r target-setup, echo=TRUE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE)

# wrangling packages
library(here)
library(janitor)
library(readxl)
library(data.table)
library(stringr)

# analysis packages
library(emmeans)
library(car) # qqplot, spreadlevel
library(DHARMa)

# graphing packages
library(ggsci)
library(ggpubr)
library(ggforce)
library(cowplot)
library(lazyWeave) #pvalstring

here <- here::here
data_path <- "data"

ggplotsci_path <- here("R", "ggplotsci.R")
source(ggplotsci_path)
```

## Data source

Data source: [ASK1 inhibits browning of white adipose tissue in obesity](https://www.nature.com/articles/s41467-020-15483-7){target="_blank"} 

This chunk assigns the path to the Excel data file for all panels of Figure 2. The data for each panel are in a single sheet in the Excel file named "Source Date_Figure 2".

```{r target-file-path}
data_folder <- "ASK1 inhibits browning of white adipose tissue in obesity"
file_name <- "41467_2020_15483_MOESM4_ESM.xlsx"
file_path <- here(data_path, data_folder, file_name)
  
fig_2_sheet <- "Source Date_Figure 2"
```

## control the color palette

```{r}
fig_2_palette <- pal_okabe_ito[5:6]
# fig_2_palette <- pal_okabe_ito
#fig_2_palette <- pal_nature_mod

```

## useful functions

A function to import longitudinal data from Fig 2

```{r target-fig2-import-in-parts}
# function to read in parts of 2b
import_fig_2_part <- function(range_2){
  fig_2_part <- read_excel(file_path,
                          sheet = fig_2_sheet,
                          range = range_2,
                          col_names = TRUE) %>%
    data.table()
  group <- colnames(fig_2_part)[1]
  setnames(fig_2_part, old = group, new = "treatment")
  fig_2_part[, treatment := as.character(treatment)] # this was read as logical
  fig_2_part[, treatment := group] # assign treatment group
  fig_2_part[, mouse_id := paste(group, .I)]
  return(fig_2_part)
}

```

```{r target-fig2-AUC}
# script to compute various area under the curves (AUC) using trapezoidal method
# le Floch's "incremental" auc substracts the baseline value from all points.
# This can create some elements with negative area if post-baseline values are less
# than baseline value.
# Some researchers "correct" this by setting any(y - ybar < 0 to zero. Don't do this.

auc <- function(x, y, method="auc", average = FALSE){
  # method = "auc", auc computed using trapezoidal calc
  # method = "iauc" is an incremental AUC of Le Floch
  # method = "pos_iauc" is a "positive" incremental AUC of Le Floch but not Wolever
  # method = "post_0_auc" is AUC of post-time0 values
  # if average then divide area by duration
  if(method=="iauc"){y <- y - y[1]}
  if(method=="pos_iauc"){y[y < 0] <- 0}
  if(method=="post_0_auc"){
    x <- x[-1]
    y <- y[-1]
  }
  n <- length(x)
  area <- 0
  for(i in 2:n){
    area <- area + (x[i] - x[i-1])*(y[i-1] + y[i])
  }
  value <- area/2
  if(average == TRUE){
    value <- value/(x[length(x)] - x[1])
  }
  return(value)
}

```

## figure 2b -- effect of ASK1 deletion on growth (body weight)
### figure 2b -- import

```{r target-fig2b-import}
range_list <- c("A21:N41", "A43:N56", "A58:N110", "A112:N170")
fig_2b_wide <- data.table(NULL)
for(range_i in range_list){
  part <- import_fig_2_part(range_i)
  fig_2b_wide <- rbind(fig_2b_wide,
                       part)
}

fig_2b <- melt(fig_2b_wide,
               id.vars = c("treatment", "mouse_id"),
               variable.name = "week",
               value.name = "body_weight")
fig_2b[, week := as.numeric(as.character(week))]
fig_2b[, c("ask1", "diet") := tstrsplit(treatment, " ", fixed=TRUE)]
fig_2b[, week_f := factor(week)]

```

### figure 2b -- exploratory plots

```{r target-fig2b-initial-plot-1}
qplot(x = week,
      y = body_weight,
      data = fig_2b,
      color = treatment) +
  facet_grid(ask1~diet)
```

1. no obvious outliers
2. reduced growth rate at bigger size

```{r target-fig2b-initial-plot-2}
qplot(x = week,
      y = body_weight,
      data = fig_2b,
      color = treatment) +
  geom_smooth()
```

1. loess smooth. Growth in ASK1F/F + HFD clearly greater than other three treatment combinations.


```{r target-fig2b-analysis}

```

## Figure 2c -- Effect of ASK1 deletion on final body weight
### Figure 2c -- import

```{r target-fig2c-import}
range_2c <- "A173:BD176"
y_cols <- c("ASK1F/F chow", "ASK1Δadipo chow", "ASK1F/F HFD", "ASK1Δadipo HFD")
fig_2c_import <- read_excel(file_path,
                     sheet = fig_2_sheet,
                     range = range_2c,
                     col_names = FALSE) %>%
  transpose(make.names=1) %>%
  data.table() %>%
  melt(measure.vars = y_cols,
       variable.name = "treatment",
       value.name = "body_weight_gain") %>%
  na.omit()
fig_2c_import[, mouse_id := paste(treatment, .I, sep = "_"),
       by = treatment]
```

### Figure 2c -- check own computation of weight change v imported value

Note that three cases are missing from fig_2c import that are in fig_2b

```{r}
# change colnames to char
fig_2c <- copy(fig_2b_wide)
weeks <- unique(fig_2b[, week])
setnames(fig_2c,
         old = colnames(fig_2c),
         new = c("treatment", paste0("week_", weeks), "mouse_id"))
fig_2c[, weight_gain := week_12 - week_0]
fig_2c <- fig_2c[, .SD, .SDcols = c("treatment",
                                "week_0",
                                "week_12",
                                "weight_gain")]
fig_2c[, mouse_id := paste(treatment, .I, sep = "_"),
       by = treatment]

fig_2c[, c("ask1", "diet") := tstrsplit(treatment, " ", fixed=TRUE)]

fig_2c_check <- merge(fig_2c,
                fig_2c_import,
                by = c("mouse_id"),
                all = TRUE)

#View(fig_2c_check)


```

### Figure 2c -- exploratory plots

```{r}
qplot(x = treatment,
      y = weight_gain,
      data = fig_2c)
```

* no obvious outliers
* variance increases with mean, as expected from growth, suggests a multiplicative model. But start with simple lm.

### Figure 2c -- fit the model: m1 (lm)

```{r}
fig_2c_m1 <- lm(weight_gain ~ week_0 + ask1*diet, data = fig_2c)
```

### Figure 2c -- check the model: m1

```{r}
# check normality assumption
set.seed(1)
qqPlot(fig_2c_m1, id=FALSE)
```

```{r}
spreadLevelPlot(fig_2c_m1, id=FALSE)

```

* QQ indicates possible right skew but especially left side is squashed toward mean
* spread-level indicates variance increases with mean

For p-value, this may not be too severe but for intervals, best to account for this. Try gamma with log link (which makes biological sense for growth)

### Figure 2c -- fit the model: m2 (gamma glm)

```{r}
fig_2c_m2 <- glm(weight_gain ~ week_0 + ask1*diet,
          family = Gamma(link = "log"),
          data = fig_2c)
```

### Figure 2c -- check the model, m2

```{r}
set.seed(1)
fig_2c_m2_sim  <-  simulateResiduals(fig_2c_m2,  n=250)
plot(fig_2c_m2_sim) 

```

* well behaved QQ and spread-level

### Figure 2c -- inference from the model
```{r, message = FALSE}
coef_table <- cbind(coef(summary(fig_2c_m2)),
                    exp(confint(fig_2c_m2))) %>%
  data.table(keep.rownames = TRUE)
coef_table[, Estimate:=exp(Estimate)]
knitr::kable(coef_table, digits = c(0,2,2,2,4,2,2))
```

```{r}
fig_2c_m2_emm <- emmeans(fig_2c_m2, specs = c("diet", "ask1"),
                         type = "response")
fig_2c_m2_pairs <- contrast(fig_2c_m2_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none") %>%
  summary(infer = TRUE)
```

```{r}
fig_2c_m2_emm
```

```{r}
fig_2c_m2_pairs
```

* within ASK1 Cn, HFD mean is 1.6X chow mean
* within ASK1 KO, HFD mean is 1.2X chow mean
* within chow, ASK1 KO mean is 1.0X ASK1 Cn mean
* within HFD, ASK1 KO mean is 0.86X ASK1 Cn mean

```{r}
# same as interaction effect in coefficient table
contrast(fig_2c_m2_emm, interaction = "pairwise", by = NULL) %>%
  summary(infer = TRUE)
```
* the reduction in weight gain in the ASK1 KO mice compared to ASK1 CN is 0.785X. Notice that p > 0.05.

### Figure 2c -- plot the model

```{r, message=FALSE}
fig_2c_m2_emm_dt <- summary(fig_2c_m2_emm) %>%
  data.table
fig_2c_m2_pairs_dt <- data.table(fig_2c_m2_pairs)
fig_2c_m2_pairs_dt[ , p_pretty := pvalString(p.value)]

dodge_width <- 0.8 # separation between groups
# get x positions of brackets for p-values
# requires looking at table and mentally figuring out
# Chow is at x = 1 and HFD is at x = 2
fig_2c_m2_pairs_dt[, group1 := c(1-dodge_width/4,
                                 1+dodge_width/4,
                                 1-dodge_width/4,
                                 2-dodge_width/4)]
fig_2c_m2_pairs_dt[, group2 := c(2-dodge_width/4,
                                 2+dodge_width/4,
                                 1+dodge_width/4,
                                 2+dodge_width/4)]

pd <- position_dodge(width = dodge_width)
fig_2c_gg <- ggplot(data = fig_2c,
                    aes(x = diet,
                        y = weight_gain,
                        color = ask1)) +
  
  # points
  geom_sina(alpha = 0.5,
            position = pd) +
  
  # plot means and CI
  geom_errorbar(data = fig_2c_m2_emm_dt,
                aes(y = response,
                    ymin = asymp.LCL,
                    ymax = asymp.UCL,
                    color = ask1),
                width = 0,
                position = pd
  ) +
  
  geom_point(data = fig_2c_m2_emm_dt,
             aes(y = response,
                 color = ask1),
             size = 3,
             position = pd
  ) +
  
  # plot p-values (y positions are adjusted by eye)
  stat_pvalue_manual(fig_2c_m2_pairs_dt,
                     label = "p_pretty",
                     y.position=c(28.5, 31, 26, 26),
                     tip.length = 0.01) +
  
  # aesthetics
  ylab("Weight Gain") +
  scale_color_manual(values=fig_2_palette,
                     name = NULL) +
  theme_pubr() +
  theme(legend.position="top") +
  theme(axis.title.x=element_blank()) +
  
  NULL

fig_2c_gg
```

### Figure 2c -- report

Results could be reported using either:

(This is inconsistent with plot, if using this, the plot should reverse what factor is on the x-axis and what factor is the grouping (color) variable) Mean weight gain in ASK1F/F mice on HFD was 1.56 (95% CI: 1.33, 1.82, $p < 0.0001$) times that of ASK1F/F mice on chow while mean weight gain in ASK1Δadipo mice on HFD was only 1.22 (95% CI: 1.01, 1.47, $p = 0.036$) times that of ASK1Δadipo mice on chow. This reduction in weight gain in ASK1Δadipo mice compared to ASK1F/F control mice was 0.79 times (95% CI; 0.61, 1.00, $p = 0.0531$).

(This is consistent with the plot in that its comparing difference in the grouping factor within each level of the factor on the x-axis) Mean weight gain in ASK1Δadipo mice on chow was trivially larger (1.03 times) than that in ASK1F/F mice on chow (95% CI: 0.83, 1.27, $p = 0.81$) while mean weight gain in ASK1Δadipo mice on HFD was smaller (0.81 times) than that in ASK1F/F control mice on HFD (95% CI: 0.72 , 0.91, $p = 0.0003$). This reduction in weight gain in ASK1Δadipo mice compared to ASK1Δadipo mice is 0.79 times (95% CI; 0.61, 1.00, $p = 0.0531$).

**note to research team**. The big difference in *p*-values between weight difference on chow and weight difference on HFD might lead one to believe there is a "difference in this difference". Using a *p*-value = effect strategy, this is not supported.

## Figure 2d -- Effect of ASK1 KO on glucose tolerance (whole curve)
### Figure 2d -- Import

```{r}
range_list <- c("A179:H189", "A191:H199", "A201:H214", "A216:H230")
fig_2d_wide <- data.table(NULL)
for(range_i in range_list){
  part <- import_fig_2_part(range_i)
  fig_2d_wide <- rbind(fig_2d_wide,
                       part)
}
fig_2d_wide[, c("ask1", "diet") := tstrsplit(treatment, " ", fixed=TRUE)]

# melt
fig_2d <- melt(fig_2d_wide,
               id.vars = c("treatment",
                           "ask1",
                           "diet",
                           "mouse_id"),
               variable.name = "time",
               value.name = "glucose")
fig_2d[, time := as.numeric(as.character(time))]

# for plot only (not analysis!)
shift <- 2
fig_2d[treatment == "ASK1F/F chow", time_x := time - shift*1.5]
fig_2d[treatment == "ASK1Δadipo chow", time_x := time - shift*.5]
fig_2d[treatment == "ASK1F/F HFD", time_x := time + shift*.5]
fig_2d[treatment == "ASK1Δadipo HFD", time_x := time + shift*1.5]

```

### Figure 2d -- exploratory plots
```{r}

qplot(x = time_x, y = glucose, color = treatment, data = fig_2d) +
  geom_line(aes(group = mouse_id), alpha = 0.3)
```
* no obvious unplausible outliers but two mice w/ high values in "F/F HFD"
* similar at time zero (initial effect is trivial)

use AUC conditional on time 0 glucose

```{r}
qplot(x = time,
      y = glucose,
      data = fig_2d,
      color = treatment) +
  geom_smooth()

```

### Figure 2d -- fit the model

### Figure 2d -- check the model

### Figure 2d -- inference

### Figure 2d -- plot the model

## Figure 2e -- Effect of ASK1 deletion on glucose tolerance (summary measure)

The researchers did create a table to import but this analysis uses the mean post-baseline glucose amount as the response instead of the area under the curve of over the full 120 minutes. This mean is computed as the post-baseline area under the curve divided by the duration of time of the post-baseline measures (105 minutes). This analysis will use fig_2d_wide since there is only one a single Y variable per mouse.

### Figure 2e -- message the data

```{r}
# AUC of post-baseline values
# do this after melt as we don't need this in long format)
fig_2e <- fig_2d_wide
fig_2e[, glucose_0 := get("0")]

times <- c(0, 15, 30, 45, 60, 90, 120)
time_cols <- as.character(times)
Y <- fig_2e[, .SD, .SDcols = time_cols]
fig_2e[, glucose_mean := apply(Y, 1, auc,
                           x=times,
                           method = "post_0_auc",
                           average = TRUE)]
```

### Figure 2e -- exploratory plots

```{r}
qplot(x = treatment, y = glucose_mean, data = fig_2e)
```

### Figure 2e -- fit the model
```{r}
fig_2e_m1 <- lm(glucose_mean ~ glucose_0 + ask1*diet, data = fig_2e)
```

### Figure 2e -- check the model

```{r}
# check normality assumption
set.seed(1)
qqPlot(fig_2e_m1, id=FALSE)
```

```{r}
spreadLevelPlot(fig_2e_m1, id=FALSE)

```
### Figure 2e -- inference from the model

```{r}
fig_2e_m1_coef <- coef(summary(fig_2e_m1))
fig_2e_m1_coef
```
```{r}
fig_2e_m1_emm <- emmeans(fig_2e_m1, specs = c("diet", "ask1"))
fig_2e_m1_pairs <- contrast(fig_2e_m1_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none") %>%
  summary(infer = TRUE)
```

```{r}
fig_2e_m1_emm
```
```{r}
fig_2e_m1_pairs
```

### Figure 2e -- plot the model
```{r, message=FALSE}
fig_2e_gg <- ggplot_the_response(
  fig_2e_m1,
  fig_2e_m1_emm,
  fig_2e_m1_pairs,
  wrap_col=NULL,
  y_label = "Post-baseline glucose (mmol per l)",
  contrast_rows = "all",
  palette = pal_okabe_ito_4,
)

fig_2e_gg

```

## Figure 2f -- Effect of ASK1 deletion  on glucose infusion rate
### Figure 2f -- import
```{r, message=FALSE}
range_2f <- "A239:I240"
treatment_levels <- c("ASK1F/F", "ASK1Δadipo")
fig_2f <- read_excel(file_path,
                     sheet = fig_2_sheet,
                     range = range_2f,
                     col_names = FALSE) %>%
  transpose(make.names=1) %>%
  data.table() %>%
  melt(measure.vars = treatment_levels,
       variable.name = "treatment",
       value.name = "glucose_infusion_rate") %>%
  na.omit()

fig_2f[, treatment := factor(treatment, treatment_levels)]
```

### Figure 2f -- exploratory plots

### Figure 2f -- fit the model

```{r}
fig_2f_m1 <- lm(glucose_infusion_rate ~ treatment, data = fig_2f)
```

### Figure 2f -- check the model

### Figure 2f -- inference

```{r}
fig_2f_m1_coef <- summary(fig_2f_m1) %>%
  coef()
fig_2f_m1_emm <- emmeans(fig_2f_m1, specs = "treatment")
fig_2f_m1_pairs <- contrast(fig_2f_m1_emm,
                            method = "revpairwise") %>%
  summary(infer = TRUE)
```

```{r}
fig_2f_m1_pairs
```

### Figure 2f -- plot the model

```{r}
fig_2f_m1_emm_dt <- summary(fig_2f_m1_emm) %>%
  data.table
fig_2f_m1_pairs_dt <- data.table(fig_2f_m1_pairs)
fig_2f_m1_pairs_dt[ , p_pretty := pvalString(p.value)]
fig_2f_m1_pairs_dt[, group1 := 1]
fig_2f_m1_pairs_dt[, group2 := 2]


fig_2f_gg <- ggplot(data = fig_2f,
                    aes(x = treatment,
                        y = glucose_infusion_rate,
                        color = treatment)) +
  
  # points
  geom_sina(alpha = 0.5) +
  
  # plot means and CI
  geom_errorbar(data = fig_2f_m1_emm_dt,
                aes(y = emmean,
                    ymin = lower.CL,
                    ymax = upper.CL,
                    color = treatment),
                width = 0
  ) +
  
  geom_point(data = fig_2f_m1_emm_dt,
             aes(y = emmean,
                 color = treatment),
             size = 3
  ) +
  
 # plot p-values (y positions are adjusted by eye)
  stat_pvalue_manual(fig_2f_m1_pairs_dt,
                     label = "p_pretty",
                     y.position=c(95),
                     tip.length = 0.01) +
  
  # aesthetics
  ylab("Glucose infusion rate") +
  scale_color_manual(values=fig_2_palette,
                     name = NULL) +
  theme_pubr() +
  theme(legend.position="none") +
  theme(axis.title.x=element_blank()) +
  
  NULL

fig_2f_gg
```

## Figure 2g -- Effect of ASK1 deletion on tissue-specific glucose uptake

### Figure 2g -- import
```{r}
range_list <- c("A244:G247", "A250:H253")
# import ASK1F/F
fig_2g_1 <- read_excel(file_path,
                     sheet = fig_2_sheet,
                     range = "A244:G247",
                     col_names = FALSE) %>%
  transpose(make.names=1) %>%
  data.table()
fig_2g_1[, treatment := "ASK1F/F"]

# import ASK1Δadipo
fig_2g_2 <- read_excel(file_path,
                     sheet = fig_2_sheet,
                     range = "A250:H253",
                     col_names = FALSE) %>%
  transpose(make.names=1) %>%
  data.table()
fig_2g_2[, treatment := "ASK1Δadipo"]

# combine
fig_2g <- rbind(fig_2g_1, fig_2g_2)


```

### Figure 2g -- exploratory plots

### Figure 2g -- fit the model

```{r}
# a more sophisticated would be a mixed model to dampen noise
fig_2g_m1_ingWAT <- lm(ingWAT ~ treatment, data = fig_2g)
fig_2g_m1_epiWAT <- lm(epiWAT ~ treatment, data = fig_2g)
fig_2g_m1_Muscle <- lm(Muscle ~ treatment, data = fig_2g)
fig_2g_m1_BAT <- lm(BAT ~ treatment, data = fig_2g)
```


### Figure 2g -- check the model

### Figure 2g -- inference
```{r}
fig_2g_infer <- function(m1){
  m1_emm <- emmeans(m1, specs = "treatment")
  m1_pairs <- contrast(m1_emm,
                              method = "revpairwise") %>%
    summary(infer = TRUE)
  return(list(emm = m1_emm,
              pairs = m1_pairs))
}

fig_2g_m1_emm_dt <- data.table(NULL)
fig_2g_m1_pairs_dt <- data.table(NULL)
m1_list <- list(fig_2g_m1_ingWAT, 
             fig_2g_m1_epiWAT,
             fig_2g_m1_Muscle,
             fig_2g_m1_BAT)
y_cols <- c("ingWAT", "epiWAT", "Muscle", "BAT")
for(i in 1:length(y_cols)){
  m1_infer <- fig_2g_infer(m1_list[[i]])

  m1_emm_dt <- summary(m1_infer$emm) %>%
    data.table
  fig_2g_m1_emm_dt <- rbind(fig_2g_m1_emm_dt,
                            data.table(tissue = y_cols[i], m1_emm_dt))
  m1_pairs_dt <- m1_infer$pairs %>%
    data.table
  fig_2g_m1_pairs_dt <- rbind(fig_2g_m1_pairs_dt,
                            data.table(tissue = y_cols[i], m1_pairs_dt))
}

```


```{r}
fig_2g_m1_pairs_dt
```

### Figure 2g -- plot the model

```{r, message = FALSE, warning=FALSE}
# melt fig_2g

fig_2g_long <- melt(fig_2g,
                    id.vars = "treatment",
                    variable.name = "tissue",
                    value.name = "glucose_uptake")

# change name of ASK1Δadipo label
fig_2g_long[treatment == "ASK1Δadipo", treatment := "ASK1-/-adipo"]
fig_2g_m1_emm_dt[treatment == "ASK1Δadipo", treatment := "ASK1-/-adipo"]

fig_2g_m1_pairs_dt[ , p_pretty := pvalString(p.value)]
fig_2g_m1_pairs_dt[, group1 := 1]
fig_2g_m1_pairs_dt[, group2 := 2]

fig_2g_plot <- function(tissue_i,
                        y_lab = FALSE, # title y-axis?
                        g_lab = FALSE  # add group label?
                        ){
  y_max <- max(fig_2g_long[tissue == tissue_i, glucose_uptake], na.rm=TRUE)
  y_min <- min(fig_2g_long[tissue == tissue_i, glucose_uptake], na.rm=TRUE)
  y_pos <- y_max + (y_max-y_min)*.05
  
  gg <- ggplot(data = fig_2g_long[tissue == tissue_i],
                      aes(x = treatment,
                          y = glucose_uptake,
                          color = treatment)) +
    
    # points
    geom_sina(alpha = 0.5) +
    
    # plot means and CI
    geom_errorbar(data = fig_2g_m1_emm_dt[tissue == tissue_i],
                  aes(y = emmean,
                      ymin = lower.CL,
                      ymax = upper.CL,
                      color = treatment),
                  width = 0
    ) +
    
    geom_point(data = fig_2g_m1_emm_dt[tissue == tissue_i],
               aes(y = emmean,
                   color = treatment),
               size = 3
    ) +
    
    # plot p-values (y positions are adjusted by eye)
    stat_pvalue_manual(fig_2g_m1_pairs_dt[tissue == tissue_i],
                       label = "p_pretty",
                       y.position=c(y_pos),
                       tip.length = 0.01) +
    
    # aesthetics
    ylab("Glucose Uptake") +
    scale_color_manual(values=fig_2_palette,
                       name = NULL) +
    ggtitle(tissue_i)+
    
    theme_pubr() +
    theme(legend.position="top") +
    theme(axis.title.x = element_blank(),
          axis.text.x = element_blank(),
          axis.ticks.x = element_blank()) +
    
    NULL
  
  if(y_lab == FALSE){
    gg <- gg + theme(axis.title.y = element_blank())
  }
  if(g_lab == FALSE){
    gg <- gg + theme(legend.position="none")
  }
  
  return(gg)
}

y_cols <- c("ingWAT", "epiWAT", "Muscle", "BAT")
legend <- get_legend(fig_2g_plot("ingWAT", y_lab = TRUE, g_lab = TRUE))
gg1 <- fig_2g_plot("ingWAT", y_lab = TRUE, )
gg2 <- fig_2g_plot("epiWAT")
gg3 <- fig_2g_plot("Muscle")
gg4 <- fig_2g_plot("BAT")
top_gg <- plot_grid(gg1, gg2, gg3, gg4,
                       nrow=1,
                       rel_widths = c(1.15, 1, 1.05, 1.1)) # by eye
plot_grid(top_gg, legend,
          nrow=2,
          rel_heights = c(1, 0.1))
```

## Figure 2h

## Figure 2i -- Effect of ASK1 deletion on liver TG

```{r}
range_2i <- "A265:G266"
treatment_levels <- c("ASK1F/F", "ASK1Δadipo")
fig_2i <- read_excel(file_path,
                     sheet = fig_2_sheet,
                     range = range_2i,
                     col_names = FALSE) %>%
  transpose(make.names=1) %>%
  data.table() %>%
  melt(measure.vars = treatment_levels,
       variable.name = "treatment",
       value.name = "liver_tg") %>%
  na.omit()

fig_2i[, treatment := factor(treatment, treatment_levels)]

# View(fig_2i)
```

### Figure 2i -- fit the model

```{r, echo = TRUE}
fig_2i_m1 <- lm(liver_tg ~ treatment, data = fig_2i)
```

### Figure 2i -- check the model

```{r, echo = TRUE}
set.seed(1)
qqPlot(fig_2i_m1, id=FALSE)
```


```{r, echo = TRUE}
spreadLevelPlot(fig_2i_m1, id=FALSE)
```

* The QQ plot looks okay, in the sense that the observed data points (open circles) fall within the boundaries set by the dashed line.
* spread level looks pretty good
* Fit normal model

### Figure 2i -- inference

```{r}
fig_2i_m1 <- lm(liver_tg ~ treatment, data = fig_2i)
fig_2i_m1_coef <- cbind(coef(summary(fig_2i_m1)),
                        confint(fig_2i_m1))
fig_2i_m1_emm <- emmeans(fig_2i_m1, specs = "treatment")
fig_2i_m1_pairs <- contrast(fig_2i_m1_emm,
                            method = "revpairwise") %>%
  summary(infer = TRUE)
```

```{r}
fig_2i_m1_pairs
```

### Figure 2i -- plot the model

```{r, echo = TRUE}

fig_2i_m1_emm_dt <- summary(fig_2i_m1_emm) %>%
  data.table
fig_2i_m1_pairs_dt <- data.table(fig_2i_m1_pairs)
fig_2i_m1_pairs_dt[ , p_pretty := pvalString(p.value)]
fig_2i_m1_pairs_dt[, group1 := 1]
fig_2i_m1_pairs_dt[, group2 := 2]


fig_2i_gg <- ggplot(data = fig_2i,
                    aes(x = treatment,
                        y = liver_tg,
                        color = treatment)) +
  
  # points
  geom_sina(alpha = 0.5) +
  
  # plot means and CI
  geom_errorbar(data = fig_2i_m1_emm_dt,
                aes(y = emmean,
                    ymin = lower.CL,
                    ymax = upper.CL,
                    color = treatment),
                width = 0
  ) +
  
  geom_point(data = fig_2i_m1_emm_dt,
             aes(y = emmean,
                 color = treatment),
             size = 3
  ) +
  
 # plot p-values (y positions are adjusted by eye)
  stat_pvalue_manual(fig_2i_m1_pairs_dt,
                     label = "p_pretty",
                     y.position=c(95),
                     tip.length = 0.01) +
  
  # aesthetics
  ylab("Liver TG (µmol per g liver)") +
  scale_color_manual(values=fig_2_palette,
                     name = NULL) +
  theme_pubr() +
  theme(legend.position="none") +
  theme(axis.title.x=element_blank()) +
  
  NULL

fig_2i_gg
```

### Figure 2i -- report the model

Mean TG level in ASK1Δadipo mice on a high-fat diet was 21.6 µmol/g less than that in ASK1F/F mice on a high-fat diet (95% CI: -37.3, -5.9, $p = 0.012$) (Figure xxx).

## Figure 2j

