# Estimating effects and uncertainty with a statistical model

1. Explore the data
2. Fit the model
3. Check the model fit
4. Estimate the predicted means
5. Estimate the contrasts between groups
6. Plot the data and model summary
7. Write the methods
8. Write the results

## Which model do I fit?

If you want to model the variance due to sampling with a normal distribution, use a *linear model* using the function `lm`.

```{r, eval = FALSE}
m1 <- lm(y ~ treatment, data = dt)
```

If you want to model the sampling variance with a specific distribution, use a *generalized linear model* using the function `glm`.

```{r, eval=FALSE}
m2 <- glm(y ~ treatment, family = binomial, data = dt)
```

If you want to model non-independence in subsets of observations that come in batches, then use a *mixed model* by adding a random factor. If you add the random factor to a linear model, this is a *linear mixed model*, which can be fit with the function `lmer`.

```{r, eval = FALSE}
# random intercept only 
m3 <- lmer(y ~ treatment + (1|batch), data = dt)
# random intercept and slope
m4 <- lmer(y ~ treatment + (treatment|batch), data = dt)
```

If you add the random factor to a generalized linear model, this is a *generalized linear mixed model*, which can be fit with the function `glmer`.

```{r, eval=FALSE}
# random intercept only 
m5 <- glmer(y ~ treatment + (treatment|batch), family = binomial, data = dt)
# random intercept and slope
m6 <- glmer(y ~ treatment + (treatment|batch), family = binomial, data = dt)
```


