---
title: "32-factorial_effects"
output: html_notebook
---

```{r type3-setup, echo=FALSE, message=FALSE}
library(car)
library(ggplot2)
library(data.table)
library(emmeans)
#source("../R/gg_interaction.R")
source("R/gg_interaction.R") # bookdown

```
# Estimating effects in factorial designs

## Create fake urchin data that simulate Fig. 2b

First, we estimate the means and standard errors from the published figure From these, compute model coefficients (using R's default dummy coding) and the error ($\sigma$).
```{r type3-fd-setup}
# modeling means and sd in Fig 2b
n <- 6
means <- c(8.2, 12.5, 7.8, 9.5) # temp-/co2-, temp+/co2-, temp-/co2+, temp+/co2+
ses <- c(0.5, 1, 0.5, 0.75) # temp-/co2-, temp+/co2-, temp-/co2+, temp+/co2+
sds <- ses*sqrt(n) # temp-/co2-, temp+/co2-, temp-/co2+, temp+/co2+
sigma <- mean(sds)
b <- c(means[1], (means[2]-means[1]), (means[3]-means[1]))
b <- c(b, means[4] - (b[1] + b[2] + b[3]))
```

Second, create fake factor columns that are common to all fake data
```{r type3-fd-factors}
temp_levels <- rep(c("Temp-", "Temp+"), each=n) # 13C, 18C
co2_levels <- c("CO2-", "CO2+") # 400 µATM, 1100 µATM
x <- data.table(expand.grid(Temp=temp_levels, CO2=co2_levels)) # columns of factor data

fd <- data.table(Resp=NA, x)
X <- model.matrix(~Temp*CO2, x)
```

This is very very close to original data.
```{r fd0}
set.seed(132)
e <- rnorm(n*4, mean=0, sd=sigma)
fd[, Resp:=X%*%b + e]
coefficients(summary(lm(Resp ~ Temp*CO2, data=fd))) # type I coefficients

```

### Main effects
#### Dummy coding -- no missing with and without interaction in model
Estimate is the same but the SE is bigger in the additive model (including interaction variance?) -- is this a necessary consequence? or, a general but not necessary consequence? ANOVA results same as contrasts.
```{r dummy-coding-no-missing}
fit <- lm(Resp ~ Temp*CO2, data=fd)
cell_means <- data.table(summary(emmeans(fit, specs=c("Temp", "CO2"))))
emmeans(fit, specs=("Temp"))
mean(cell_means[Temp=="Temp-", emmean])
mean(cell_means[Temp=="Temp+", emmean])
contrast(emmeans(fit, specs=("Temp")), method="revpairwise", adjust="none")
contrast(emmeans(fit, specs=("CO2")), method="revpairwise", adjust="none")
anova(fit)
Anova(fit, type=2)
X.dummy <- model.matrix(fit)

fit2 <- lm(Resp ~ Temp+CO2, data=fd)
cell_means2 <- data.table(summary(emmeans(fit2, specs=c("Temp", "CO2"))))
emmeans(fit2, specs=("Temp"))
mean(cell_means2[Temp=="Temp-", emmean])
mean(cell_means2[Temp=="Temp+", emmean])
contrast(emmeans(fit2, specs=("Temp")), method="revpairwise", adjust="none")
contrast(emmeans(fit2, specs=("CO2")), method="revpairwise", adjust="none")
anova(fit2)
Anova(fit2, type=2)

```

#### Dummy coding -- missing with and without interaction in model
Estimates for type factorial and additive models differ in addition to SEs differing.
```{r dummy_coding-missing}
exc <- which(fd[,Temp]=="Temp+" & fd[, CO2]=="CO2+")
inc <- setdiff((1:nrow(fd)), exc[1])
fit <- lm(Resp ~ Temp*CO2, data=fd[inc])
cell_means <- data.table(summary(emmeans(fit, specs=c("Temp", "CO2"))))
emmeans(fit, specs=("Temp"))
mean(cell_means[Temp=="Temp-", emmean])
mean(cell_means[Temp=="Temp+", emmean])
contrast(emmeans(fit, specs=("Temp")), method="revpairwise", adjust="none")
contrast(emmeans(fit, specs=("CO2")), method="revpairwise", adjust="none")
anova(fit)
Anova(fit, type="2")
con3 <- list(Temp=contr.sum, CO2=contr.sum)
Anova(lm(Resp~Temp*CO2, data=fd[inc]), type="3")


fit2 <- lm(Resp ~ Temp+CO2, data=fd[inc])
cell_means2 <- data.table(summary(emmeans(fit2, specs=c("Temp", "CO2"))))
emmeans(fit2, specs=("Temp"))
mean(cell_means2[Temp=="Temp-", emmean])
mean(cell_means2[Temp=="Temp+", emmean])
contrast(emmeans(fit2, specs=("Temp")), method="revpairwise", adjust="none")
contrast(emmeans(fit2, specs=("CO2")), method="revpairwise", adjust="none")
anova(fit2)
Anova(fit2, type="2")

```

#### Effects coding -- missing with and without interaction in model
This is the same as with dummy coding with missing

```{r effects-coding-missing}
con3 <- list(Temp=contr.sum, CO2=contr.sum)
fit <- lm(Resp ~ Temp*CO2, data=fd[inc], contrasts=con3)
cell_means <- data.table(summary(emmeans(fit, specs=c("Temp", "CO2"))))
emmeans(fit, specs=("Temp"))
mean(cell_means[Temp=="Temp-", emmean])
mean(cell_means[Temp=="Temp+", emmean])
contrast(emmeans(fit, specs=("Temp")), method="revpairwise", adjust="none")
X.effects <- model.matrix(fit)

fit2 <- lm(Resp ~ Temp+CO2, data=fd[inc], contrasts=con3)
cell_means2 <- data.table(summary(emmeans(fit2, specs=c("Temp", "CO2"))))
emmeans(fit2, specs=("Temp"))
mean(cell_means2[Temp=="Temp-", emmean])
mean(cell_means2[Temp=="Temp+", emmean])
contrast(emmeans(fit2, specs=("Temp")), method="revpairwise", adjust="none")

```

#### simulation
b[4] = -2.6 or nearly 2sigma
with no missing, full model has smaller se 83% of the runs, average SE 0.678 (full) vs. 0.736 (add)
with missing, full model has closer estimate 53% of runs, RMSE 0.708 (full) vs. 0.712 (add)

b[4] = .32 or 2sigma
with missing, full model has smaller se 35% of the runs, average SE 0.694 (full) vs. 0.695 (add)
with missing, full model has closer estimate 52% of runs, RMSE 0.692 (full) vs. 0.691 (add)

```{r simulate-non-missing}
# same parameters as above but I'm just hard coding it in here
n <- 6
means <- c(8.2, 12.5, 7.8, 9.5) # temp-/co2-, temp+/co2-, temp-/co2+, temp+/co2+
ses <- c(0.5, 1, 0.5, 0.75) # temp-/co2-, temp+/co2-, temp-/co2+, temp+/co2+
sds <- ses*sqrt(n) # temp-/co2-, temp+/co2-, temp-/co2+, temp+/co2+
sigma <- mean(sds)
b <- c(means[1], (means[2]-means[1]), (means[3]-means[1]))
b <- c(b, means[4] - (b[1] + b[2] + b[3])) # b4 will change in simulation

b[4] <- .32 # =.2sigma
temp_levels <- rep(c("Temp-", "Temp+"), each=n) # 13C, 18C
co2_levels <- c("CO2-", "CO2+") # 400 µATM, 1100 µATM
x <- data.table(expand.grid(Temp=temp_levels, CO2=co2_levels)) # columns of factor data

fd <- data.table(Resp=NA, x)
X <- model.matrix(~Temp*CO2, x)

niter <- 1000
theta1 <- (b[1] + (b[1] + b[3]))/2 # expected marginal mean of temp-
theta2 <- ((b[1] + b[2]) + (b[1] + b[2] + b[3] + b[4]))/2 # expected marginal mean of temp+
theta <- theta2 - theta1 # expected effect
d.full <- numeric(niter)
d.add <- numeric(niter)
se.full <- numeric(niter)
se.add <- numeric(niter)

do_missing <- TRUE
exc <- which(fd[,Temp]=="Temp+" & fd[, CO2]=="CO2+")
full_inc <- 1:24
missing_inc <- setdiff(full_inc, exc[1])
inc <- ifelse(do_missing==TRUE, missing_inc, full_inc) # why doesn't this work?
if(do_missing==TRUE){inc <- missing_inc}else{inc <- full_inc}

for(iter in 1:niter){
  e <- rnorm(n*4, mean=0, sd=sigma)
  fd[, Resp:=X%*%b + e]
  fit1 <- lm(Resp ~ Temp*CO2, data=fd[inc])
  emm1 <- emmeans(fit1, specs=("Temp"))
  d.full[iter] <- summary(contrast(emm1, method="revpairwise", adjust="none"))[1, "estimate"]
  se.full[iter] <- summary(contrast(emm1, method="revpairwise", adjust="none"))[1, "SE"]
  fit2 <- lm(Resp ~ Temp+CO2, data=fd[inc])
  emm2 <- emmeans(fit2, specs=("Temp"))
  d.add[iter] <- summary(contrast(emm2, method="revpairwise", adjust="none"))[1, "estimate"]
  se.add[iter] <- summary(contrast(emm2, method="revpairwise", adjust="none"))[1, "SE"]
}
sqrt(mean((d.full-theta)^2))
sqrt(mean((d.add-theta)^2))
sum((abs(d.full-theta) - abs(d.add-theta)) < 0)/niter
sd(d.full-theta)
sd(d.add-theta)
mean(se.full)
mean(se.add)
mean(se.full-se.add)
sd(se.full-se.add)
sum(se.full < se.add)/niter
```

### What are the coefficients estimating in the full model?

1. Interaction is significant ($p=0.026$)
2. CO2 main effect is significant ($p=0.004$)
3. CO2 at Temp- is not significant ($p=0.57$)

The results and plot are like the original (Fig. 2b) except that the interaction is significant here.

```{r fd1}
set.seed(1)
e <- rnorm(n*4, mean=0, sd=sigma)
fd[, Resp:=X%*%b + e]
coefficients(summary(lm(Resp ~ Temp*CO2, data=fd))) # type I coefficients

# type 3 anova
con3 <- list(Temp=contr.sum, CO2=contr.sum)
fit3 <- lm(Resp ~ Temp*CO2, data=fd, contrasts=con3)
coef.full <- coefficients(summary(fit3))
coef.full
Anova(fit3, type="3")

# what are the type 3 coefficients
# b1 is the difference between the grand mean and the pooled mean of X1+ across both levels of X2
b.full <- coef.full[, "Estimate"]
mt <- means.table(x=c("Temp","CO2"), y="Resp", data=fd)
b0 <- mt[3,3] # compare to b_check
temp.pos.mm <- mt["Temp+", "Temp-mm"]
CO2.pos.mm <- mt["CO2-mm", "CO2+"]
b1 <- b0 - temp.pos.mm # compare to b_check
b2 <- b0 - CO2.pos.mm # compare to b_check

# interaction plot with meaning of type III coefficients
# using means table
gg <- gg_interaction_plot(x=c("Temp","CO2"), y="Resp", data=fd) +
  geom_hline(yintercept=b0, linetype="dashed") +
  geom_hline(yintercept=CO2.pos.mm, color="red", linetype="dashed") + # co2 effect
  geom_hline(yintercept=temp.pos.mm, color="blue", linetype="dashed") + #temp effect
  NULL
gg

```

### What are the coefficients estimating in the additive model?

```{r fd1-additive}
fit3 <- lm(Resp ~ Temp + CO2, data=fd, contrasts=con3)
coef.add <- coefficients(summary(fit3))
coef.add
Anova(fit3, type="3")

# what are the type 3 coefficients
# b1 is the difference between the grand mean and the pooled mean of X1+ across both levels of X2
b.add <- coef.add[, "Estimate"]
mt <- means.table(x=c("Temp","CO2"), y="Resp", data=fd)
b0 <- mt[3,3] # compare to b_check
temp.pos.mm <- mt["Temp+", "Temp-mm"]
CO2.pos.mm <- mt["CO2-mm", "CO2+"]
b1 <- b0 - temp.pos.mm # compare to b_check
b2 <- b0 - CO2.pos.mm # compare to b_check

```

## fake data with intermediate CO2 effect at Temp- but big CO2 effect at Temp+

The idea is to create a negative CO2 effect at both levels of Temp and then ask "what does a CO2 main effect mean?"

```{r}
# need to make b3 a bit bigger
b1 <- copy(b)
delta <- -1.8
b1[3] <- b1[3] + delta
fd[, Resp:=X%*%b1 + e]
coefficients(summary(lm(Resp ~ Temp*CO2, data=fd)))
con3 <- list(Temp=contr.sum, CO2=contr.sum)
fit3 <- lm(Resp ~ Temp*CO2, data=fd, contrasts=con3)
coef_table <- coefficients(summary(fit3))
coef_table
Anova(fit3, type="3")

grand_mean <- coef_table["(Intercept)", "Estimate"]
co2pos_mean <- grand_mean - coef_table["CO21", "Estimate"]
gg <- gg_interaction_plot(x=c("Temp","CO2"), y="Resp", data=fd) +
  geom_hline(yintercept=grand_mean, linetype="dashed") +
  geom_hline(yintercept=co2pos_mean, color="red", linetype="dashed") +
  NULL
gg

```

