---
title: "Models   two (or more) categorical X variables -- Factorial designs"
author: "Jeffrey A. Walker"
date: "11/13/2020"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    fig_caption: yes
    number_sections: true
    code_folding: show
    includes:
      before_body: copyright.html
---
# Linear models with two categorical $X$ -- Factorial linear models ("two-way ANOVA") {#factorial}

```{r twoway-setup, echo=FALSE, warning=FALSE, message=FALSE}
library(here)
library(janitor)
library(readxl)
library(data.table)

# analysis packages
library(emmeans)
library(car) # qqplot, spreadlevel

# graphing an tabling packages
library(ggplot2) # ggplot environment
library(ggpubr) # publication ready plots
library(cowplot) # combine plots
library(knitr)
library(kableExtra) #tables
library(equatiomatic)

ggplot_the_model_path <- here::here("R/ggplot_the_model.R")
source(ggplot_the_model_path)

here <- here::here
clean_names <- janitor::clean_names
data_folder <- "data"
minus <- "\u2013"
```

## A linear model with crossed factors estimates interaction effects

A factorial experiment is one in which there are two or more factor variables (categorical $X$) that are **crossed**, resulting in a group for each combination of the levels of each factor. Each specific combination is a different treatment. A **linear model with crossed factors** is used to estimate **interaction effects**, which occur when the effect of the level of one factor is conditional on the level of the other factors. Estimation of the interaction effect is necessary for inferences about

1. "Something different" -- Estimation of a treatment effect relative to a control effect (Example 1 -- TLR9-/- mice)
2.  "It depends" -- Estimation of the effect of background condition on an effect (Example 2 -- XX mice)
3. "More than the sum of the parts" -- Estimation of **synergy**, a non-additive effect (Example 3 -- plant root growth)

Inferences like these are common in the experimental biology literature but they are made using the wrong statistics. The correct statistic -- the interaction effect -- is easy to compute but rarely computed.

<div style="background-color:#cccccc; text-align:left; vertical-align: middle; padding:20px 47px;">
Alert. NHST encourages thinking such as "there is or isn't an interaction effect" This dichotomous thinking can only lead to disasters of biblical proportion, dogs and cats living together. Interaction effects in biology are ubiquitous. Sometimes they are small enough to ignore.
</div>

### An interaction is a difference in simple effects

In this chapter, I'll describe an interaction effect using different descriptions (Examples 1-3). Here, I want to emphasize that an interaction effect is a difference of differences. To clarify this, I'll introduce a fake experiment. A research group has evidence that a certain gene product (CGP) is an intermediary between intestinal microbiota and obesity. The researchers transfer feces from either lean mice or obese mice into either wildtype mice or CGP-/- mice to investigate the effect of microbiota and CGP on body weight. The design has two factors (donor and genotype), each with two levels (donor: "Lean", "Obese; genotype: "WT", "KO"), which makes this a $2 \times 2$ (crossed or factorial) design. There are $n=6$ replicates for each treatment.

A good way to visualize the treatment combinations in a crossed design is with a $m \times p$ table showing all combinations of the $m$ levels of factor 1 ($\texttt{donor}$) against the $p$ levels of factor 2 ($\texttt{genotype}$) (\@ref(tab:twoway-factorial-table)).

```{r twoway-factorial-table, echo=FALSE}
row_names <- c("Lean", "Obese", "Effect")
col_names <- c("WT", "KO", "Effect")
wt_col <- c(33.2, 41.9)
wt_col <- c(wt_col, wt_col[2] - wt_col[1])
ko_col <- c(34.1, 32.7)
ko_col <- c(ko_col, ko_col[2] - ko_col[1])
diff_col <- c(ko_col[1] - wt_col[1],
              ko_col[2] - wt_col[2],
              (ko_col[2] - wt_col[2]) - (ko_col[1] - wt_col[1]))

dat <- matrix(c(wt_col,
                ko_col,
                diff_col), nrow = 3)
colnames(dat) <- col_names
dt <- data.table(
  " " = row_names,
  dat
)
dt %>%
  kable(caption = "Mean body weight of mice in four treatment groups of the 2 x 2 factorial experiment.") %>%
  kable_paper("hover", full_width = F) %>%
  column_spec(1, bold = T) %>%
  column_spec(4, bold = T,
              color = c("#696969", "#696969", "#FF0000")) %>%
  row_spec(3, bold = T)

```

1. The upper-left $2 \times 2$ part of the table contains the fake mean body weight of each treatment group. These means are known as **cell means**.
2. The first two elements in the "Effect" column contains the difference of the two cells to the left -- these are the **effects of genotype conditional on the level of donor**.
3. The first two elements in the "Effect" row contains the difference of the two cells above -- these are the **effects of donor conditional on the level of genotype**.
4. The effects described in items 2 and 3 are known as the **simple effects**.
5. The value in red is the difference of the two simple effects above it. It is also the difference of the two simple effects to the left. These differences are equal. This is the interaction effect.

In this fake experiment, we want to know the effect of obese donor treatment in the KO mice *compared to* the effect of obese donor treatment in the WT mice. That is, we want the contrast of these two simple effects.

$$
(\operatorname{Obese/KO} - \operatorname{Lean/KO}) - (\operatorname{Obese/WT} - \operatorname{Lean/WT})
$$

This contrast is the interaction effect. **An interaction effect is a difference of differences**.

### A linear model with crossed factors includes interaction effects

The factorial linear model for the fake data is

$$
\begin{align}
\texttt{body_weight} &= \beta_0 + \beta_1 (\texttt{donor}_{\texttt{Obese}}) + \beta_2 (\texttt{genotype}_{\texttt{KO}})\ + \\
&\quad \ \beta_3 (\texttt{donor}_{\texttt{Obese}} : \texttt{genotype}_{\texttt{KO}}) + \varepsilon
\end{align}
$$
What are the variables?

1. $\texttt{donor}_{\texttt{Obese}}$ is the indicator variable for the "Obese" level of the factor $\texttt{donor}$. It contains the value 1 if the donor was "Obese" and 0 otherwise.
2. $\texttt{genotype}_{\texttt{KO}}$ is the indicator variable for the "KO" level of the factor $\texttt{genotype}$. It contains the value 1 if the genotype is "KO" and 0 otherwise.
3. $\texttt{donor}_{\texttt{Obese}} : \texttt{genotype}_{\texttt{KO}}$ is the indicator variable for the interaction between the "Obese" level of $\texttt{donor}$ and the "KO" level of $\texttt{genotype}$. This ":" character to indicate interaction follows the R formula convention in the LME4 package. Many sources use a $\times$ symbol instead of the colon. The variable contains the value 1 if the mouse is assigned to both "Obese" and "KO" and 0 otherwise. This value is the product of the values in $\texttt{donor}_{\texttt{Obese}}$ and $\texttt{genotype}_{\texttt{KO}}$.

What are the parameters?

This linear model has set "Lean" as the reference level of $\texttt{donor}$ and "WT" as the reference level of $\texttt{genotype}$. This make the Lean/WT mice the control.

1. $\beta_0$ is the true mean of the control, which is the "Lean/WT" group.
2. $\beta_1$ is the true effect of donor in the WT mice (the effect of manipulating the donor factor but not the genotype factor). It is the difference between the true means of the "Obese/WT" group and the "Lean/WT" group.

The mean of the "Obese/WT" group is $\beta_0 + \beta_1$. This is the expectation if we start with the control and then add the effect of Obese donor.

3. $\beta_2$ is the true effect of genotype in the mice given feces from lean donors  (the effect of manipulating the genotype factor but not the donor factor). It is the difference between the true means of the "Lean/KO" group and the "Lean/WT" group.

The mean of the "Lean/KO" group is $\beta_0 + \beta_2$. This is the expectation if we start with the control and then add the effect of KO genotype.

4. $\beta_3$ is the true interaction effect of $\texttt{donor}_{\texttt{Obese}} : \texttt{genotype}_{\texttt{KO}}$

The mean of the Obese/KO group is $\beta_0 + \beta_1 + \beta_2 + \beta_3$. The expected mean of the Obese/KO group, *if the factors are additive, which means the interaction effect is zero*, is $\beta_0 + \beta_1 + \beta_2$. **The interaction effect is the difference between the actual mean of the Obese/KO group and this additive mean**. The interaction effect is what's left-over, after you've added Obese effect and the KO effect to the control.

### factorial experiments are frequently analyzed as flattened linear models in the experimental biology literature

Often, researchers analyze data from a factorial experiment with a one-way ANOVA followed by pairwise tests (or by a simple series of separate *t*-tests). For the fake experiment, this linear model is

$$
\begin{align}
\texttt{body_weight} &= \beta_0 + \beta_1 (\texttt{treatment}_{\texttt{Obese}}) + \beta_2 (\texttt{treatment}_{\texttt{KO}}) \ + \\
& \quad \ \beta_3 (\texttt{treatment}_{\texttt{Obese + KO}}) + \varepsilon
\end{align}
$$

I refer to this as a **flattened** model (the table of treatment combinations has been flattened into a single row). Pairwise tests between treatments from a factorial analysis and a flattened analysis are the same. What differs is the interaction effect is not estimated in the flattened analysis (look at the two equations). This is a critical difference. The interaction effect is the necessary statistic to make many of the claims reported in the experimental biology literature.

## Example 1 -- Estimation of a treatment effect relative to a control effect ("Something different")  (TLR9-/- mice)

To introduce a linear model with crossed factors (categorical $X$ variables), I'll use data from a set of experiments designed to measure the effect of the toll-like receptor protein TLR9 on the activation of excercise-induced AMP-activated protein kinase (AMPK) and downstream sequelae of this activation, including glucose transport (from outside to inside the cell) by skeletal muscle cells.

Article source: [TLR9 and beclin 1 crosstalk regulates muscle AMPK activation in exercise](https://www.nature.com/articles/s41586-020-1992-7){target="_blank"}

[Public source](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC7047589/){target="_blank"}

The data are from multiple experiments in Figure 2.

[Data source](https://www.nature.com/articles/s41586-020-1992-7#Sec27){target="_blank"}

```{r twoway-exp2j-import, echo=FALSE}
data_from <- "TLR9 and beclin 1 crosstalk regulates muscle AMPK activation in exercise"
file_name <- "41586_2020_1992_MOESM4_ESM.xlsx"

file_path <- here(data_folder, data_from, file_name)

treatment_levels  <- c("WT Rest",
                       "WT Stimulation",
                       "KO Rest",
                       "KO Stimulation")
exp2j_wide <- read_excel(file_path,
                         sheet = "2j",
                         range = "A5:D13",
                         col_names = TRUE) %>%
  data.table()

colnames(exp2j_wide) <- treatment_levels

exp2j <- melt(exp2j_wide,
              measure.vars = treatment_levels,
              variable.name = "treatment",
              value.name = "glucose_uptake") %>%
  na.omit() # danger!

exp2j[, c("genotype", "stimulation") := tstrsplit(treatment,
                                                  " ",
                                                  fixed = TRUE)]

genotype_levels <- c("WT", "KO")
stimulation_levels <- c("Rest", "Stimulation")
exp2j[, genotype := factor(genotype,
                           levels = genotype_levels)]
exp2j[, stimulation := factor(stimulation,
                              levels = stimulation_levels)]
# View(exp2j)
```

### Understand the experimental design

**Background**. Exercise (muscle activity) stimulates AMPK activated glucose uptake (transport from outside to inside the muscle cell). Instead of natural exercise, which can stimulate multiple systems, the researchers directly activate the muscle with electrical stimulation.

**Research question** There are two ways to think about the question -- both produce the same answer.

1. How much does TLR9 knockout inhibit the expected increase of glucose uptake following electrical stimulation? The expected increase comes from the contrast of the positive and negative controls. Call this the knockout-induced stimulation effect. 
2. How much does TLR9 knockout inhibit glucose uptake during muscle stimulation compared to effect of TLR9 knockuot during muscle rest? Call this the **stimulation-induced TLR9 effect**. 

**Response variable** $\texttt{glucose_uptake}$ (nmol per mg protein per 15 min) -- the rate of glucose transported into the cell.

**Factor 1** -- $\texttt{genotype}$ ("WT", "KO").

* "WT" (reference level) -- C57BL/6J mice with intact TLR gene (TLR+/+)
* "KO" -- TLR-/- mice on a C57BL/6J background

**Factor 2** $\texttt{activity}$ ("Rest", "Stimulation") -- Two levels:

* "Rest" (reference level) -- muscle that has not been stimulated.
* "Stimulation" -- electrical stimulation of muscle to induce contraction and contractile-related cell changes

These two factors create the three control treatments and the one focal treatment:

* WT/Rest -- **Negative control**. Expect non-exercise (low) level uptake.

* WT/Stimulation -- **Positive control**. Expect high level uptake.

* KO/Rest -- **Method control**. Unsure of KO effect on uptake, which is why we need this control.

* KO/Stimulation -- **Focal treatment**. At about same level of KO/Rest if TLR9-/- completely inhibits electrical stimulation of glucose uptake

**Design** -- $2 \times 2$, that is, two crossed factors each with two levels. This results in four groups, each with a unique combination of the levels from each factor.

**Planned Contrasts**

The two ways of framing the research question suggest either of the following sets of contrasts. Both ways of framing generate a **treatment** contrast that includes the focal treatment and a **control contrast** that does not include the focal treatment. The question pursued by the experiment is addressed with the contrast of the treatment and control contrasts, which is is the estimate of the interaction effect. There is no difference in inference between the two ways of framing the question -- the interaction effects are equivalent. The two framings simply give two different ways of viewing an interaction effect.

**Set 1**: How much does TLR9 knockout inhibit the expected increase of glucose uptake during stimulation?

1. (KO/Stimulation - KO/Rest) -- effect of Stimulation in KO mice. This is our **treatment contrast**. If TLR9 is necessary for glucose uptake, then this should be zero. If positive, there are non-TLR9 paths.
 
2. (WT/Stimulation - WT/Rest) -- This is the **control (or reference) contrast** -- it is what we know based on prior knowledge and what we want to compare the treatment effect to. This should be positive based on prior knowledge.

We need to control for the expected increase in 2 using the contrast:

3. (KO/Stimulation - KO/Rest) - (WT/Stimulation - WT/Rest) -- This contrast is the interaction effect. This **focal effect** is what we need to estimate the knockout-induced stimulation effect.

**Set 2**: What is stimulation-induced TLR9 effect?

1. (KO/Stimulation - WT/Stimulation) -- effect of KO during muscle stimulation. This is our **treatment effect**. If TLR9 is necessary for glucose uptake, then this should be big and negative.
 
2. (KO/Rest - WT/Rest) -- effect of KO when the muscle is not stimulated. This is the **control (or reference) effect**. If TLR9-/- KO has non-muscle-stimulation paths to glucose uptake, this will be something other than zero.

We need to control for any (KO/Rest - WT/Rest) effect using the contrast:

3. (KO/Stimulation - WT/Stimulation) - (KO/Rest - WT/Rest) -- This contrast is the interaction effect. This **focal effect** is what we need to estimate the stimulation-induced TLR9 effect.

### Fit the linear model

```{r twoway-exp2j_m1, echo=TRUE}
exp2j_m1 <- lm(glucose_uptake ~ stimulation * genotype, data = exp2j)

```

### Inference

**The coefficient table**
```{r twoway-exp2j_m1_coef, echo=TRUE}
exp2j_m1_coef <- cbind(coef(summary(exp2j_m1)),
                         confint(exp2j_m1))
exp2j_m1_coef %>%
  kable(digits = c(2,3,1,4,2,2)) %>%
  kable_styling()
```
**emmeans table**

```{r twoway-exp2j_m1_emm, echo=TRUE}
exp2j_m1_emm <- emmeans(exp2j_m1,
                          specs = c("genotype", "stimulation"))
exp2j_m1_emm %>%
  kable(digits = c(1,1,2,3,1,2,2)) %>%
  kable_styling()
  
```

**The contrasts table**

```{r twoway-exp2j_m1_planned, echo=TRUE}
# exp2j_m1_emm # print in console to get row numbers
# set the mean as the row number from the emmeans table
wt_rest <- c(1,0,0,0)
ko_rest <- c(0,1,0,0)
wt_stim <- c(0,0,1,0)
ko_stim <- c(0,0,0,1)

# contrasts are the difference in the vectors created above
# these planned contrasts are described above
# 1. (KO/Stimulation - KO/Rest) # treatment
# 2. (WT/Stimulation - WT/Rest) # control
# 3. (KO/Stimulation - KO/Rest) - (WT/Stimulation - WT/Rest) Interaction

exp2j_m1_planned <- contrast(
  exp2j_m1_emm,
  method = list(
    "(KO/Stim - KO/Rest)" = c(ko_stim - ko_rest),
    "(WT/Stim - WT/Rest)" = c(wt_stim - wt_rest)
  ),
  adjust = "none"
) %>%
  summary(infer = TRUE)

exp2j_m1_interaction <- contrast(exp2j_m1_emm,
                           interaction = c("revpairwise"),
                           by = NULL) %>%
  summary(infer = TRUE)

# Make the interaction table conform (same column names)
# to the planned table and  remove cols 1:2
exp2j_m1_interaction <- cbind(contrast = "interaction",
                              exp2j_m1_interaction[, -c(1:2)])
# combine
exp2j_m1_planned <- rbind(exp2j_m1_planned,
                          exp2j_m1_interaction)

exp2j_m1_planned %>%
  kable(digits = c(0,3,4,0,3,3,2,5)) %>%
  kable_styling()

# double check with automated contrasts
# contrast(exp2b.2_m1_emm, method = c("revpairwise"))

```

### Plot the model

```{r twoway-exp2j-plot, echo = TRUE}
ggplot_the_model(
  exp2j_m1,
  exp2j_m1_emm,
  exp2j_m1_planned,
  palette = pal_okabe_ito_blue,
  legend_position = "bottom",
#  y_label = expression(paste("Glucose uptake (", mm^2, ")")),
  y_label = expression(paste("Glucose uptake (nmol mg ", protein^-1, "15 ", min^-1, ")")),
  effect_label = expression(paste("Effect (nmol mg ", protein^-1, "15 ", min^-1, ")")),
  contrast_rows = "all"
)

```
Alternative plot

```{r twoway-exp2j-altplot, echo=TRUE, eval=FALSE}
ggplot_the_response(
  exp2j_m1,
  exp2j_m1_emm,
  exp2j_m1_planned,
  palette = pal_okabe_ito_blue,
  legend_position = "bottom",
  y_label = "p-TBC1D1/Total p-TBC1D1 (normalized)",
  contrast_rows = 1:2
)
```

## Understanding the linear model with crossed factors 1 (Why the interaction effect is necessary in experiments with distinct treatment and control effects)

Researchers in experimental biology report almost exclusively the *p*-values of the simple effects (Table \@ref(tab:twoway-factorial-table)) in data from factorial experiments. These can be computed even if the data are flattened into a single "treatment" factor. But it is the interaction effect that is necessary to make many of the inferences made by the researchers. To see why, we need to understand what the coefficients in the coefficient table are.

### What the coefficients are

To understand the coefficients, it helps to use the means in the emmeans table to construct a factorial table of the cell means

```{r twoway-exp2j-cellmeans, echo=FALSE}
col_names <- levels(exp2j$stimulation)
row_names <- levels(exp2j$genotype)
values <- summary(exp2j_m1_emm)$emmean
labels <- as.character(round(values,2))
labels <- round(values,2)
dt <- data.table(
    " " = row_names,
    Rest = labels[1:2],
    Stimulation = labels[3:4]
)


dt %>%
  kable("html", escape = F,
        digits = 2,
        caption = "Cell mean table") %>%
  kable_paper("hover", full_width = F) %>%
  column_spec(1, bold = T)

```

The linear model fit to the $\texttt{exp2j}$ data is

$$
\begin{align}
\texttt{glucose_uptake} &= \beta_{0} + \beta_{1}(\texttt{genotype}_{\texttt{KO}}) + \beta_{2}(\texttt{stimulation}_{\texttt{Stimulation}}) \ + \\
&\quad \ \beta_{3}(\texttt{genotype}_{\texttt{KO}} : \texttt{stimulation}_{\texttt{Stimulation}}) + \varepsilon\
\end{align}
$$

and the fit coefficients are

```{r twoway-exp2j_m1_coef-explainer, echo=FALSE}
exp2j_m1_coef %>%
  kable(digits = c(3,4,2,4,3,3)) %>%
  kable_styling()  
```

Explainer

1. Understand what the rows of the coefficient table are. There are four parameters in the fit linear model -- the rows are the statistics for the estimates of these parameters. These estimates are the **coefficients** of the model.
2. The $\texttt{(Intercept)}$ ($b_0$) is the mean $\texttt{glucose_uptake}$ of the reference level, which was set to the WT/Rest group (Figure \@ref(fig:twoway-what-coefs-are)). This is the mean in the upper left cell in Table \@ref(tab:twoway-exp2j-cellmeans).
3. The $\texttt{genotypeKO}$ coefficient ($b_1$) is the estimate of the **the added effect of knocking out TLR9 when the stimulation factor is at its reference level**, and so is the mean in the lower left cell minus the mean in the upper left cell in Table \@ref(tab:twoway-exp2j-cellmeans). This coefficient is not a mean -- it is a difference of means (Figure \@ref(fig:twoway-what-coefs-are)). The mean of the KO/Rest group is $b_0 + b_1$.
4. The $\texttt{stimulationStimulation}$ coefficient ($b_2$) is the estimate of the **the added effect of Stimulation when the genotype factor is at its reference level**, and so is the mean in the upper right cell minus the mean in the upper left cell in Table \@ref(tab:twoway-exp2j-cellmeans). This coefficient is not a mean -- it is a difference of means (Figure \@ref(fig:twoway-what-coefs-are)). The mean of the WT/Stimulation group is $b_0 + b_2$.
5. The $\texttt{genotypeKO:stimulationStimulation}$ coefficient ($b_3$) is the estimate of the interaction effect between $\texttt{genotype}$ and $\texttt{stimulation}$. It is *not* the mean in the lower right cell minus the mean in the upper left cell. Instead, **it is the mean in the lower right cell minus the expected mean in the lower right cell if the genotype and activity treatment effects were additive**. The expected additive-mean in the lower right cell (KO/Stimulation) is $b_0 + b_1 + b_2$. The mean of KO/stimulation is $b_0 + b_1 + b_2 + b_3$ (Figure \@ref(fig:twoway-what-coefs-are)).
6. The interaction effect is a **non-additive effect**. Think about this. Adding the "Stimulation" alone adds 3.45 nmol glucose per protein per 15 min to the uptake rate of the reference. Adding "KO" alone adds 0.78 glucose per protein per 15 min to the uptake rate of the reference. If these effects were purely additive, then adding both Stimulation and KO to the reference rate should result in a mean of 6.75 + 3.45 + 0.78 = 10.98 glucose per protein per 15 min. The modeled mean for KO/Stimulation is 8.67 glucose per protein per 15 min. The difference observed - additive is 8.67 - 10.98 = -2.31 glucose per protein per 15 min. Compare this difference to the interaction coefficient in the coefficient table.
7. Reinforce your understanding of "non-additive" in item 5. The interaction is a non-additive effect because the mean of the combined treatment is something different than if we were to just add the KO and Stimulation effects. But this effect is additive in the linear model. This is what linear models are -- a reference mean plus the sum of a bunch of effects.
8. Understand what these rows **are not**. The $\texttt{stimulationStimulation}$ row is not the same as the "stimulation" term in a Type III ANOVA table (the ANOVA table produced in GraphPad Prism or JMP). The *p*-values will be different because the *p*-values are testing different hypotheses. In the coefficient table, the $\texttt{stimulationStimulation}$ *p*-value is testing the difference of the means of WT/Stimulation and WT/Rest. The stimulation term in a Type III ANOVA table is testing if there is an overall stimulation effect, which is estimated as the average of the two stimulation contrasts (WT/Stimulation - WT/Rest) and (KO/Stimulation - KO/Rest). The average of these two contrasts is not often of interest (but sometimes is -- see below).

```{r twoway-what-coefs-are-build, echo=FALSE, eval=FALSE}
exp2j_m1_emm_dt <- summary(exp2j_m1_emm) %>%
  data.table
b <- coef(exp2j_m1)

fudge <- 0.0
nudge <- -0.05
bracket_table <- data.table(
  x = c(1 + nudge,
        2 - nudge,
        2 - nudge),
  y = c(b[1],
        b[1],
        b[1] + b[2] + b[3] + fudge),
  xend = c(1 + nudge,
           2 - nudge,
           2 - nudge),
  yend = c(b[1] + b[2],
           b[1] + b[3] - fudge,
           b[1] + b[2] + b[3] + b[4])
)

dodge_width <- 0.0
pd <- position_dodge(dodge_width)
gg_ixn_explainer_base <- ggplot(data = exp2j_m1_emm_dt,
             aes(x = genotype,
                 y = emmean,
                 color = stimulation)) +
  
  geom_hline(yintercept = b[1],
             linetype = "dashed",
             color = "black",
             alpha = 0.5) +
  
  geom_point(size = 3,
             position = pd) +
  scale_color_manual(values = pal_okabe_ito_blue) +
#  scale_x_discrete(labels = c("WT", "TLR9-/-")) +
  ylab(expression(paste("Glucose uptake (nmol mg ", protein^-1, "15 ", min^-1, ")"))) +
  coord_cartesian(ylim = c(5.5, 11.5)) +
  
  geom_line(aes(group = stimulation),
            position = pd) +
  
  # additive line
  geom_segment(aes(
    x = 1 + dodge_width/4,
    y = b[1] + b[2],
    xend = 2 + dodge_width/4,
    yend = b[1] + b[2] + b[3]),
    color = pal_okabe_ito_blue[2],
    linetype = "dashed",
    alpha = 0.2
  ) +

  # additive point
  geom_point(aes(
    x = 2 + dodge_width/4,
    y = b[1] + b[2] + b[3]),
    size = 3,
    color = pal_okabe_ito_blue[2],
    alpha = 0.1) +
  
  theme_pubr() +
  theme(axis.title.x = element_blank(),
        legend.title = element_blank()) +
  
  NULL


gg_ixn_explainer <- gg_ixn_explainer_base +

  # additive annotation
  annotate(
    geom = "text",
    x = c(2.07, 2.07),
    y = c(b[1] + b[2] + b[3],
          b[1] + b[2] + b[3] + b[4]),
    label = c("Expected mean of\nKO/Stimulation if\ninteraction is 0.0",
              "mean of\nKO/Stimulation"),
    hjust = 0,
    size = 3,
    color = "darkgray"
  ) +

  # coefficient bracket
  geom_segment(data = bracket_table,
    aes(x = x,
    y = y,
    xend = xend,
    yend = yend),
    color = "black"
  ) +

  # coefficient labels
  annotate(
    geom = "text",
    x = c(.52, .9, 2.1, 2.1),
    y = c(b[1] + 0.2,
          b[1] + 0.5*b[2],
          b[1] + 0.5*b[3],
          b[1] + b[2] + b[3] + .5*b[4]),
    label = c("b[0]",
              "b[1]",
              "b[2]",
              "b[3]"),
    parse = TRUE,
  ) +

  # b0 arrow
  geom_segment(aes(x = 0.65,
                   y = b[1],
                   xend = 0.4,
                   yend = b[1]),
               color = "black",
               arrow = arrow(length  =  unit(0.05, "npc"),
                             ends = "last",type = "open")) +

  NULL

image_path <- here("images", "twoway-what_coefs_are.png")
ggsave(image_path,
       width = 7,
       height = 5,
       units = "in")

gg_ixn_explainer

```

```{r twoway-what-coefs-are, echo=FALSE, out.width="80%", fig.cap = "The coefficients of a linear model with two crossed factors, explained."}

image_path <- here("images", "twoway-what_coefs_are.png")

knitr::include_graphics(image_path)
```

### The interaction effect is something different

Item 6 in the coefficient table explainer stated "The interaction is a non-additive effect because the mean of the combined treatment is **something different** than if we were to just add the KO and Stimulation effects. The something different is the interaction effect. If the interaction effect were zero, the expected effect of stimulation in the KO mice would be the same as the expected effect of stimulation in the WT mice (Figure \@ref(fig:twoway-something-different). This would suggest the underlying physiological changes between Rest and Stimulation in the KO mice is "more of the same" physiological changes in the WT mice. But, because of the interaction, the underlying physiological changes between Rest and Stimulation in the KO mice is "something different" to that of physiological changes in the WT mice (Figure \@ref(fig:twoway-something-different)).

The biological reasons causing interaction effects are highly variable and are what makes Biology fun. Additive effects (no interaction) may occur when combined treatments act independently of each other. This might occur in the glucose uptake response if knocking out TLR9 opens a path to glucose uptake that is different from and independent of the paths activated by electrial stimulation. Positive, or **synergistic** interaction effects may occur when combined treatments augment each other's ability to affect the response (see Example 3 below). This could occur in the glucose uptake response if knocking out TLR9 opens a path to glucose uptake that is different the paths activated by electrial stimulation but also makes the paths activated by stimulation more sensitive to stimulation. Negative, or **antagonistic** interaction effects may occur when combined treatments interfere with each other's ability to affect the response. This could occur in the glucose uptake response if TLR9 is on the path from stimulation to glucose uptake. Knocking out TLR9 interferes with this path. If TLR9 is required, we'd expect the interaction effect to be the same magnitude but opposite sign of the control effect -- that is, complete antagonism of the control effect. Previous experiments suggested this negative interaction. The measurement of this effect was the purpose of experiment $\texttt{exp2j}$.

```{r twoway-something-different-build, echo=FALSE, eval=FALSE}
b <- coef(exp2j_m1)
nudge <- 0.05
x_pos <- c(1- nudge,
           2 + 2*nudge,
           2 + nudge,
           2 - nudge,
           2 + 2*nudge)
bracket_table <- data.table(
  x = x_pos,
  y = c(b[1],
        b[1],
        b[1] + b[3], # additive diff
        b[1] + b[3], # actual diff
        b[1] + b[2] + b[3]), # something diff
  xend = x_pos,
  yend = c(b[1] + b[2], # b2
           b[1] + b[3], # b3
           b[1] + b[2] + b[3], # additive diff
           b[1] + b[2] + b[3] + b[4], # actual diff
           b[1] + b[2] + b[3] + b[4]) # something diff
)

x_pos_text <- x_pos + c(-.45, 0.02, 0.02, -0.3, 0.02)

gg_something_different <- gg_ixn_explainer_base +
    # coefficient bracket
  geom_segment(data = bracket_table,
    aes(x = x,
    y = y,
    xend = xend,
    yend = yend),
    color = "black"
  ) +

  #  labels
  annotate(
    geom = "text",
    x = x_pos_text,
    y = c(
      b[1] + 0.5*b[2],
      b[1] + 0.5*b[3],
      b[1] + b[3] + 0.5*(sum(b) - sum(b[c(1,3)])),
      b[1] + b[3] + 0.5*(sum(b) - sum(b[c(1,3)])),
      b[1] + b[2] + b[3] + .5*b[4]),
    label = c(
      "1. what stimulation\ndoes in WT",
      "2. what KO does\nin Rest",
      "3. expected difference\nif more of 1 & 2",
      "4. observed\ndifference",
      "5. something\ndifferent"),
    hjust = 0
  ) +
  
  NULL

image_path <- here("images", "twoway-something-different.png")
ggsave(image_path,
       width = 7,
       height = 5,
       units = "in")


gg_something_different
```

```{r twoway-something-different, echo=FALSE, out.width="80%", fig.cap = "The interaction is something different, not more of the same."}
image_path <- here("images", "twoway-something-different.png")

knitr::include_graphics(image_path)
```

### Why we want to compare the treatment effect to a control effect

The purpose of the experiment is to infer a TLR9 role in the regulation of muscle-stimulated glucose transport, that is, a **stimulation-induced TLR9 effect**. If TLR9 is in the pathway from muscle activity to glucose transport, we expect some kind of recovery to baseline (Rest) values in the KO/Stimulation group. But TLR9 may also (or alternatively) have a role in non-stimulation-induced glucose uptake.

How do we make an inference about a stimulation-induced TLR9 effect using this experiment? Researchers typically look at the **treatment effect**

(KO/Stimulation - WT/Stimulation)

and conclude a stimulation-induced TLR9 effect if it's big (and negative)

If the treatment effect is the correct contrast for inference, why bother with the measures of glucose uptake during Rest? The reason the treatment effect alone is the wrong contrast for inferring the *stimulation-induced* TLR9 effect is because it is confounded by the **control effect**, which is the contrast (KO/Rest - WT/Rest) (Figure \@ref(fig:twoway-what-we-want)).

```{r twoway-what-we-want-build, echo=FALSE, eval=FALSE}
b <- coef(exp2j_m1)
nudge <- 0.04
x_pos <- c(2 + nudge,
           2, # wrong diff
           2 + nudge)
bracket_table <- data.table(
  x = x_pos,
  y = c(b[1],
        b[1] + b[2] + b[3] + b[4], # wrong diff
        b[1] + b[2] + b[3]), # something diff
  xend = x_pos,
  yend = c(b[1] + b[3], # b3
           b[1] + b[2], # wrong diff
           b[1] + b[2] + b[3] + b[4]) # something diff
)

x_pos_text <- x_pos + c(0.02, -0.39, 0.02)

gg_what_we_want <- gg_ixn_explainer_base +
  
  #  bracket
  geom_segment(data = bracket_table,
    aes(x = x,
    y = y,
    xend = xend,
    yend = yend),
    color = "black"
  ) +

  #  labels
  annotate(
    geom = "text",
    x = x_pos_text,
    y = c(
      b[1] + 0.5*b[3],
      b[1] + 0.85*b[2],
      b[1] + b[2] + b[3] + .4*b[4]),
    label = c(
      "1. what KO does\nin Rest (confounder)",
      "2. wrong contrast\n(confounded by 1)",
      "3. correct contrast\n(unconfounded by 1)"),
    hjust = 0,
    size = 3.5
  ) +
  
  # b2 baseline
  geom_segment(x = 1,
               y = b[1]+b[2],
               xend = 2,
               yend = b[1]+b[2],
               linetype = "dashed",
               color = "gray") +
  
  NULL

image_path <- here("images", "twoway-what-we-want.png")
ggsave(image_path,
       width = 7,
       height = 5,
       units = "in")

gg_what_we_want
```

```{r twoway-what-we-want, echo=FALSE, out.width="80%", fig.cap = "Why the interaction effect is the stimulation-induced TLR9 effect"}
image_path <- here("images", "twoway-what-we-want.png")

knitr::include_graphics(image_path)
```

To unconfound the inference, subtract the confounder:

(KO/Stimulation - WT/Stimulation) - (KO/Rest - WT/Rest)

This is the interaction effect. Consequences of interpreting the treatment effect KO/Stimulation - WT/Stimulation as the stimulation-induced TLR9 effect are highlighted in Figure \@ref(fig:twoway-scenarios). In these plots, the data are those from $\texttt{exp2j}$ but with the values in the KO groups shifted up or down to create the scenarios.

Scenario 1. The positive control has big effect AND KO has no effect during rest AND the stimulation-induced TLR9 effect is equal in magnitude but opposite direction to the Stimulation effect in WT (Figure \@ref(fig:twoway-scenarios)A). The stimulation-induced TLR9 effect is conspicuous from the plot, if our sample means are close to the true means and we have high precision. The simple effect measures the stimulation-induced TLR9 effect correctly -- but this is because the  simple effect is equal in magnitude (but opposite sign) to the interaction effect. Many experiments in the literature are pretty similar to this scenario.

Scenario 2. The positive control has big effect AND KO has a negative effect during rest BUT there is zero stimulation-induced TLR9 effect -- that is, the interaction is zero (Figure \@ref(fig:twoway-scenarios)B). There is an effect of KO during Stimulation but this effect is no different that that occuring during Rest. So this cannot be a *contraction induced* TLR9 effect. Simple effect 1 *inflates* the effect.

Scenario 3. The positive control has big effect AND KO has a positive effect during rest AND there is a stimulation-induced TLR9 effect that is equal to that in scenario 1 (Figure \@ref(fig:twoway-scenarios)C). The positive effect of KO during Rest masks the stimulation-induced TLR9 effect. The simple effect *underestimates* the stimulation-induced TLR9 effect. This is similar to what is happening in Experiment 2j.

```{r twoway-scenarios-build, echo = FALSE, eval=FALSE}

gg_plot_it <- function(fit,
                       fit_emm,
                       fit_pairs){
  b <- coef(fit)

  x_pos <- c(2.5, 3)
  y_pos <- c(b[1] + b[3], b[1] + b[2] + b[3])
  
  bracket_table <- data.table(
    x = x_pos,
    y = y_pos,
    xend = x_pos,
    yend = c(b[1] + b[2] + b[3] + b[4], # wrong diff
             b[1] + b[2] + b[3] + b[4]) # something diff
  )
  
  x_pos_text <- x_pos + c(-0.43, 0.02)
  
  gg <- ggplot_the_response(fit,
                            fit_emm,
                            fit_pairs,
                            y_label = "glucose uptake",
                            g_label = "none",
                            palette = pal_okabe_ito_blue,
                            contrast_rows = "none") +
    coord_cartesian(xlim = c(0.5, 3)) +
    
    #  bracket
    geom_segment(data = bracket_table,
                 aes(x = x,
                     y = y,
                     xend = xend,
                     yend = yend),
                 color = "black"
    ) +
    
    annotate(geom = "text",
             x = x_pos,
             y = c(max(y_pos) + .2, max(y_pos) + .2),
             label = c("Simple", "Interaction"),
             angle = 90,
             hjust = 0)
    
    NULL
  return(gg)
}

fake_exp2j <- copy(exp2j)

# observed
fake_exp2j[, scen0 := glucose_uptake]
m0 <- lm(scen0 ~ genotype * stimulation, data = fake_exp2j)
m0_emm <- emmeans(m0, specs = c("genotype", "stimulation"))
m0_pairs <- contrast(m0_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none")
gg0 <- gg_plot_it(m0, m0_emm, m0_pairs)


b <- exp2j_m1_coef[, "Estimate"]

# scenario 1 - ixn = - effect
fake_exp2j[, scen1 := glucose_uptake]
fake_exp2j[treatment == "KO Rest", scen1 := scen1 - b[3]]
fake_exp2j[treatment == "KO Stimulation", scen1 := scen1 - (b[2]+b[3]+b[4])]

m1 <- lm(scen1 ~ genotype * stimulation, data = fake_exp2j)
m1_emm <- emmeans(m1, specs = c("genotype", "stimulation"))
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none")
gg1 <- gg_plot_it(m1, m1_emm, m1_pairs)

# scenario 2 - ixn = 0 with negative effect
b2 <- 1
b3 <- -0.5
fake_exp2j[, scen2 := glucose_uptake]
fake_exp2j[treatment == "WT Stimulation",
           scen2 := scen2 - (1-b2)*b[2]]
fake_exp2j[treatment == "KO Rest",
           scen2 := scen2 - b[3] + b3*b2*b[2]]
fake_exp2j[treatment == "KO Stimulation",
           scen2 := scen2 - (b[2]+b[3]+b[4]) + b3*b2*b[2] + b[2]]

m2 <- lm(scen2 ~ genotype * stimulation, data = fake_exp2j)
m2_emm <- emmeans(m2, specs = c("genotype", "stimulation"))
m2_pairs <- contrast(m2_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none")
gg2 <- gg_plot_it(m2, m2_emm, m2_pairs)

# scenario 3 - ixn = 1/2 effect
b2 <- 1
b3 <- 0.5 # relative to b2
fake_exp2j[, scen3 := glucose_uptake]
fake_exp2j[treatment == "WT Stimulation",
           scen3 := scen3 - (1-b2)*b[2]]
fake_exp2j[treatment == "KO Rest",
           scen3 := scen3 - b[3] + b3*b2*b[2]]
fake_exp2j[treatment == "KO Stimulation",
           scen3 := scen3 - (b[2]+b[3]+b[4]) + b3*b2*b[2]]

m3 <- lm(scen3 ~ genotype * stimulation, data = fake_exp2j)
m3_emm <- emmeans(m3, specs = c("genotype", "stimulation"))
m3_pairs <- contrast(m3_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none")
gg3 <- gg_plot_it(m3, m3_emm, m3_pairs)

gg_scenarios <- plot_grid(gg1, gg2, gg3, ncol = 3, labels = "AUTO")

image_path <- here("images", "twoway-scenarios.png")
# ggsave(image_path,
#        width = 7,
#        height = 5,
#        units = "in")

png(image_path,
    width = 800,
    height = 400,
    units = "px") 
  gg_scenarios
dev.off() 

gg_scenarios

```

```{r twoway-scenarios, echo=FALSE, out.width="100%", fig.cap = "Scenarios to show the consequence of inferring the stimulation-induced TLR9 effect from the simple effect (KO/Stimulation - WT/Stimulation). The simple effect and interaction effect lines extend from the KO/Stimulation mean to either the KO/Rest mean (simple) or the expected mean of KO/Stimulation if the two factors were additive (interaction)."}
image_path <- here("images", "twoway-scenarios.png")

knitr::include_graphics(image_path)
```

## Example 2: Estimation of the effect of background condition on an effect ("it depends") (XX mice)

```{r twoway-exp3e-import, message=FALSE, echo=FALSE}

data_from <- "XX sex chromosome complement promotes atherosclerosis in mice"
file_name <- "41467_2019_10462_MOESM6_ESM.xlsx"
file_path <- here(data_folder, data_from, file_name)

# fig 3e
exp3e_wide <- read_excel(file_path,
                  sheet = "Figure 3E",
                  range = "A3:K4",
                  col_names = FALSE) %>%
  data.table() %>%
  transpose(make.names = 1)

sex_levels <- colnames(exp3e_wide)
exp3e <- melt(exp3e_wide,
              measure.vars = sex_levels,
              variable.name = "sex",
              value.name = "lesian_area")

# convert lesian_area to mm^2 from µm^2
exp3e[, lesian_area := lesian_area/10^6]

# convert sex variable to factor
exp3e[, sex := factor(sex,
                        levels = sex_levels)]

# create chromosome column and convert to factor
chromosome_levels <- c("XX", "XY")
exp3e[, chromosome := rep(rep(chromosome_levels, each = 5), 2)]
exp3e[, chromosome := factor(chromosome,
                             levels = chromosome_levels)]

# researchers treatment levels
exp3e[, treatment := rep(c("FXX", "FXY", "MXX", "MXY"), each = 5)]

# two rows with missing response so delte these rows as there is 
# no useful information in them
exp3e <- na.omit(exp3e) # be careful with a global na.omit

# View(exp3e) # highlight and run to see data
```

### Understand the experimental design

**Research question** Does the X complement stimulate dyslipidemia and lipid-related diseases and, if so, how conditional is the effect on background gonad?

**Response variable** $\texttt{lesian_area}$ -- atherosclerotic lesian area in aortic sinus.

**Factor 1** -- $\texttt{sex}$ ("Female", "Male"). This is not the typical sex factor that is merely observed but is a manipulated factor. Sex is determined by the presence or absence of *SRY* on an autosome using the Four Core Genotype mouse model. *SRY* determines the gonad that develops (ovary or testis). "Female" does not have the autosome with *SRY*. "Male" has the autosome with *SRY*.

**Factor 2** -- $\texttt{chromosome}$ ("XX", "XY"). The sex chromosome complement is not observed but manipulated. In "XX", neither sex chromosome has *SRY* as the natural condition. In "XY", *SRY* has been removed from the Y chromosome.

**Design** -- $2 \times 2$, that is, two crossed factors each with two levels. This results in four groups, each with a unique combination of the levels from each factor. "Female/XX" is the control. "Male/XX" adds the autosomal *SRY* gene (and testes instead of an ovary). "Female/XY" replaces the "X" complement with the engineered Y complement. "Male/XY" has both the autosomal *SRY* and the engineered Y complement.

```{r two-way-factor-table, echo=FALSE}
factor_table <- data.table(
  sex = c("Female", "Male", "Female", "Male"),
  chromosome = c("XX", "XX", "XY", "XY"),
  treatment = c("FXX", "MXX", "FXY", "MXY"),
  "chromosome complement" = c("X", "X", "Y (sry-)", "Y (sry-)"),
  autosome = c("WT", "SRY+", "WT", "SRY+"),
  gonad = c("ovary", "testis", "ovary", "testis")
)

knitr::kable(factor_table) %>%
  kable_styling()
```

The research question suggest the following contrasts

Does the X complement stimulate dyslipidemia and lipid-related diseases?

1. (Female/XX - Female/XY) -- effect of XX in mice with female gonad. If hypothesis is true, this should be large, negative.
2. (Male/XX - Male/XY) -- effect of XX in mice with male gonad. If hypothesis is true, this should be large, negative.

If so, how conditional is the effect on sex?

3. (Male/XX - Male/XY) - (Female/XX - Female/XY) -- Interaction.

In this experiment, there is no treatment contrast or control contrast. Instead, there are three contrasts of equal interest -- the effect of XX in female mice, the effect of XX in male mice, and the difference in these effects.

### Fit the linear model

```{r two-way-lm, echo=TRUE}
exp3e_m1 <- lm(lesian_area ~ sex*chromosome, data = exp3e)
```

### Check the model

```{r, echo=TRUE}
ggcheck_the_model(exp3e_m1)
```

### Inference from the model

**The coefficient table**

```{r two-way-coef, echo=TRUE}
# step 2 - get the coefficient table
exp3e_m1_coef <- cbind(coef(summary(exp3e_m1)),
                 confint(exp3e_m1))
exp3e_m1_coef %>%
  kable(digits = c(3,4,2,4,3,3)) %>%
  kable_styling()
```

**The emmeans table**

```{r two-way-emm, echo=TRUE}
# step 3 - get the modeled means
exp3e_m1_emm <- emmeans(exp3e_m1, specs = c("sex", "chromosome"))

exp3e_m1_emm %>%
  summary() %>%
  kable(digits = c(1,1,3,4,1,3,3)) %>%
  kable_styling()
```

**The contrasts table**

```{r two-way-planned, echo=TRUE}
# m1_emm # print in console to get row numbers
# set the mean as the row number from the emmeans table
fxx <- c(1,0,0,0)
mxx <- c(0,1,0,0)
fxy <- c(0,0,1,0)
mxy <- c(0,0,0,1)

# contrasts are the difference in the vectors created above
# the focal contrasts are in the understand the experiment section
# 1. (FXY - FXX) 
# 2. (MXY - MXX)
# 3. Interaction

exp3e_m1_planned <- contrast(exp3e_m1_emm,
                       method = list(
                         "FXY - FXX" = c(fxy - fxx),
                         "MXY - MXX" = c(mxy - mxx)
                       ),
                       adjust = "none"
) %>%
  summary(infer = TRUE)

exp3e_m1_interaction <- contrast(exp3e_m1_emm,
                           interaction = c("revpairwise"),
                           by = NULL) %>%
  summary(infer = TRUE)

# Make the interaction table conform (same column names)
# to the planned table
exp3e_m1_interaction <- cbind(contrast = "interaction",
                        exp3e_m1_interaction[, -c(1:2)]) # remove cols 1:2
# combine
exp3e_m1_planned <- rbind(exp3e_m1_planned, exp3e_m1_interaction)

exp3e_m1_planned %>%
  kable(digits = c(0,3,4,0,3,3,2,5)) %>%
  kable_styling()

```

Explainer

1. The two simple effects and the interaction are computed separately. If we want to adjust for three comparisons, I would use the Holm method.
2. The magnitude of the estimated effect of XX in mice with male gonads is about 1/2 that in mice with female gonads. This difference is the magnitude of the interaction.
3. Don't infer "no effect" given the p-value of the interaction. The estimate of the interaction effect has the same magnitude as the estimated effect of XX in mice with male gonads and about 1/2 the magnitude as the estimated effect of XX in mice with female gonads. The interaction *p*-value suggests caution in our confidence of the sign of this effect.

**All simple effects**

```{r two-way-simple}
# step 4 - get the contrasts
exp3e_m1_simple <- contrast(exp3e_m1_emm,
                        method = "revpairwise",
                        adjust = "none",
                        simple = "each",
                        combine = TRUE) %>%
  summary(infer = TRUE)

exp3e_m1_simple %>%
  kable(digits = c(0,1,1,3,4,1,3,3,2,5)) %>%
  kable_styling()
```

1. With $k=4$ groups, there are $k(k-1)/2$ pairwise comparisons. For many factorial designs, the comparisons of interest are the **simple effects**, which are the differences among the levels of factor 1 at each level of factor 2. For example, the difference between XY and XX means when sex is "female" (row 3 in the table) and the difference between XY and XX means when sex is "male". For a $2 \times 2$ design like this, there are four simple effects. I limited the output of the `contrast` function to the four simple effects using the `simple = "each"` and `combine = TRUE` arguments. To get all six effects, simply delete or comment out the `simple =` and `combine =` arguments.

### Plot the model

```{r plot_the_model}

exp3e_m1_plot <- ggplot_the_model(
  fit = exp3e_m1,
  fit_emm = exp3e_m1_emm,
  fit_pairs = exp3e_m1_planned[1:3,],
  palette = pal_okabe_ito_blue,
  legend_position = "bottom",
  y_label = expression(paste("Lesian area (", mm^2, ")")),
  effect_label = expression(paste("Effect (", mm^2, ")")),
  contrast_rows = "all"
)
exp3e_m1_plot
```
## Understanding the linear model with crossed factors 2 (When to care about conditional effects and when to simplify the model)

### Conditional and marginal means

```{r factorial-conditional-means, echo=FALSE, message=FALSE}
options(knitr.kable.NA = '')
ycols <- c("group", names(m1_emm))

m1_emm_dt <- emmeans(m1, specs=c("hac", "indole")) %>%
  summary() %>%
  data.table()

m1_emm_hac <- emmeans(m1, specs=c("hac")) %>%
  summary() %>%
  data.table()

m1_emm_indole <- emmeans(m1, specs=c("indole")) %>%
  summary() %>%
  data.table()


hac_0 <- c(m1_emm_dt[hac == "HAC-", emmean],
           m1_emm_hac[hac == "HAC-", emmean])
hac_1 <- c(m1_emm_dt[hac == "HAC+", emmean],
           m1_emm_hac[hac == "HAC+", emmean])
indole_means <- c(m1_emm_indole[, emmean], NA)

means_table <- data.frame("HAC-" = hac_0,
                          "HAC+" = hac_1,
                          "mean" = indole_means)
row.names(means_table) <- c("Indole-", "Indole+","mean")


bgc1 <- pal_okabe_ito[1]
bgc2 <- pal_okabe_ito[2]

knitr::kable(means_table, 
             digits = 3,
             col.names = c("HAC-", "HAC+", "mean"),
             caption = "Conditional (white background) and marginal (color background) means from full factorial model fit to ABA data") %>%
  kable_styling(full_width = FALSE) %>%
  # row_spec(row = 0,
  #          bold = F) %>%
  column_spec(column = 1,
              bold = c(T, T, T)) %>%
  column_spec(column = 4,
              background = c(bgc1, bgc1, "white")) %>%
  column_spec(column = 3,
              background = c("white", "white", bgc2)) %>%
  column_spec(column = 2, 
              background = c("white", "white", bgc2))
 

```

The conditional means from model \@ref(eq:factorial-full) are shown in the upper left $2 \times 2$ block (white background) of Table \@ref(tab:factorial-conditional-means). These means are conditional on the level of $hac$ and $indole$. The values in the last row and column are the **marginal means**, which are the means of the associated row or column cells (these values are in the margins of the table). More generally, *marginal* refers to a statistic averaged across multiple levels of another variable. The marginal means of $indole$ (orange background) are the means of the `Indole+` and `Indole-` rows. The marginal means of $hac$ (blue background) are the means of the `HAC+` and `HAC-` columns.

### Simple (conditional) effects

In a factorial experiment with crossed A and B factors, there is an effect of factor A (relative to the reference, or another level of factor A) for each of the *p* levels of factor B. And, there is an effect of factor B (relative to the reference, or another level of factor B) for each of the *m* levels of factor A.

For the maize defense response experiment, there is an Indole effect at `hac = HAC-` and a different effect at `hac = HAC+`. Similarly, there is a HAC effect at `indole = Indole-` and a different effect at `indole = Indole+`. These effects, or **contrasts** (differences in modeled means), are called the **simple effects**. I prefer **conditional effects**, since the value of the effect is conditional on the level factor of the other factor.

```{r, echo=FALSE}
m1_emm <- emmeans(m1, specs=c("hac", "indole"))
m1_pairs <- contrast(m1_emm,
                        method = "revpairwise",
                        adjust = "none",
                        simple = "each",
                        combine = TRUE)
knitr::kable(m1_pairs, 
             digits=c(1,1,1,1,2,1,1,5),
             caption = "Conditional (or \"simple\") effects of the factorial model.") %>%
  kable_styling()
```

The first two rows are the conditional effects of $hac$ in each of the levels of $Indole$. The last two rows are the conditional effects of $Indole$ in each of the levels of $hac$.

```{r factorial-simple-effects, echo = FALSE}
options(knitr.kable.NA = '')
ycols <- c("group", names(m1_emm))

m1_emm <- emmeans(m1, specs=c("hac", "indole"))
m1_pairs_dt <- contrast(m1_emm,
                        method = "revpairwise",
                        adjust = "none",
                        simple = "each",
                        combine = TRUE) %>%
  summary() %>%
  data.table()

hac_0 <- c(m1_emm_dt[hac == "HAC-", emmean],
           m1_pairs_dt[hac == "HAC-", estimate])
hac_1 <- c(m1_emm_dt[hac == "HAC+", emmean],
           m1_pairs_dt[hac == "HAC+", estimate])
hac_diffs <- c(m1_pairs_dt[1:2, estimate], NA)

simple_table <- data.frame("HAC-" = hac_0,
                          "HAC+" = hac_1,
                          "simple" = hac_diffs)
row.names(simple_table) <- c("Indole-", "Indole+","simple")



bgc1 <- pal_okabe_ito[1]
bgc2 <- pal_okabe_ito[2]

knitr::kable(simple_table, 
             digits = 3,
             col.names = c("HAC-", "HAC+", "simple"),
             caption = "Conditonal (\"simple\") effects of HAC (orange background) and Indole (blue background) on ABA concentration. The conditional means of each combination of the factor levels are in the cells with the white background. The simple effect is the difference in the means of the associated row or column.") %>%
  kable_styling(full_width = FALSE) %>%
  # row_spec(row = 0,
  #          bold = F) %>%
  column_spec(column = 1,
              bold = c(T, T, T)) %>%
  column_spec(column = 4,
              background = c(bgc1, bgc1, "white")) %>%
  column_spec(column = 3,
              background = c("white", "white", bgc2)) %>%
  column_spec(column = 2, 
              background = c("white", "white", bgc2))
 


```

One way to understand conditional effects is by using the $m \times p$ table of treatment combination means (Table \@ref(factorial-simple-effects)). The values in the right-side column (orange) are the conditional effects of HAC at each level of Indole. These values are the difference of the means in the associated row. For example, the conditional effect of $hac$ when `indole = Indole-` is 13.297 (first value in orange column). The values in the bottom row (blue) are the conditional effects of Indole at each level of HAC. These values are the difference of the means in the associated column. For example, the conditional effect of $indole$ when `hac = HAC+` is 21.594 (second value in blue row). Note that the first conditional effect for each factor has a corresponding row in the table of coefficients of the fit model.

### Marginal effects

The average of the conditional effects for a factor are the **marginal effects**, or the **main effects** in ANOVA terminology.

```{r factorial-main, echo=FALSE, warning=FALSE, message=FALSE}

m1_hac_dt <- emmeans(m1, specs = "hac") %>%
  contrast(
    method = "revpairwise") %>%
  summary() %>%
  data.table()

m1_indole_dt <- emmeans(m1, specs = "indole") %>%
  contrast(
    method = "revpairwise") %>%
  summary() %>%
  data.table()

marginal_table <- copy(simple_table)

marginal_table <- rbind(marginal_table, marginal =
                        c(NA, NA, m1_hac_dt[1, estimate]))
marginal_table <- cbind(marginal_table,
                        marginal = c(NA, NA, m1_indole_dt[1, estimate], NA))

orange_color <- pal_okabe_ito[1]
blue_color <- pal_okabe_ito[2]
gray_color <- ggsci::pal_jco(alpha = 0.6)(9)[3]

knitr::kable(marginal_table, 
             digits = 3,
             col.names = c("HAC-", "HAC+", "simple", "marginal"),
             caption = "Marginal effects of HAC (orange) and Indole (blue) on ABA concentration. Simple effects are in grey cells. The conditional means of each combination of the factor levels are in the white cells.") %>%
  kable_styling(full_width = FALSE) %>%
  # row_spec(row = 0,
  #          bold = F) %>%
  column_spec(column = 1,
              bold = c(T, T, T,T)) %>%
  column_spec(column = 2, 
              background = c("white", "white", gray_color, "white")) %>%
  column_spec(column = 3,
              background = c("white", "white", gray_color, "white")) %>%
  column_spec(column = 4,
              background = c(gray_color, gray_color, "white", orange_color)) %>%
  column_spec(column = 5,
              background = c("white", "white", blue_color, "white"))

```

<div style="background-color:#cccccc; text-align:left; vertical-align: middle; padding:20px 47px;">
Be careful with reporting marginal effects. In this example, the simple effects of HAC are very different for the `Indole-` and `Indole+` treatment levels and the mean of these two effects may not be especially relevant to the question that we are pursuing. If the simple effects of a factor are similar, then it may make more sense to report the marginal effect.
</div> 

### The additive model

Marginal effects can be useful for summarizing a general trend, but, like any average, might not be especially meaningful if there is large heterogeneity of the simple effects, which occurs when the interaction effect is large. The maize ABA response is a good example of marginal effects that would be highly misleading to present without further comment.

If an interaction effect is small, then it can be useful to estimate the effects of the two factors as if the interaction were equal to zero.

\begin{equation}
aba = \beta_0 + \beta_1 hac_{HAC^+} + \beta_2 indole_{Indole^+} + \varepsilon
(\#eq:factorial-reduced)
\end{equation}

This is a **reduced model** because one of the terms has been removed from the model. This particular reduced model is often referred to as the **additive model**, since it excludes the interaction term, which is a *product* of other terms. In R, this model is

```{r}
m2 <- lm(aba ~ hac + indole, data = fig1)
```

The model coefficients of the additive model are

```{r factorial-reduced-model, echo=FALSE}
m2_coef <- cbind(coef(summary(m2)),
        confint(m2))
knitr::kable(m2_coef, 
             digits=c(2,2,1,3),
             caption = "Model coefficients of additive model.") %>%
  kable_styling()

```

The conditional effects of the reduced model are

```{r, echo=FALSE}
m2_emm <- emmeans(m2, specs=c("hac", "indole"))
m2_pairs <- contrast(m2_emm,
                 method = "revpairwise",
                 adjust = "none",
                 simple = "each",
                 combine = TRUE) %>%
  summary(infer = TRUE)


knitr::kable(m2_pairs,
             digits = c(1,1,1,5,2,1,2,2,2,5),
             caption = "Conditional effects of additive model.") %>%
  kable_styling()

```

The table shows that, in an additive model, all conditional effects for one factor are the same for each level of the other factor. This makes sense. If the model fit is additive, the interaction effect is set to zero by the model and there cannot be differences in conditional effects among the contrasts at each of the levels of the other factor (otherwise, there would be an interaction). Also note that the conditional effects xxx. Probably a better way of thinking about this is, it doesn't make sense to compute or discuss conditional effects in an additive model. Instead, an additive model automatically computes marginal effects.

Compare the table of marginal effects of the additive model to the table of marginal effects of the full model. The estimates are the same but the *t*-values and *p*-values differ because of different degrees of freedom (the full model estimates one more parameter, the interaction effect). The estimate is the same only if the design is balanced, which means that each combination of treatment levels has the same sample size *n*.

### Reduce models for the right reason

Unless one factor truly has no effect, there will always be an interaction. As stated above, interactions are ubiquitous. If an interaction is small, it can make sense to drop the interaction term and re-fit an additive model to estimate marginal effects in order to present a simplified picture of what is going on, with the recognition that these estimates are smoothing over the heterogenity in conditional (simple) effects that truly exist.

Aided and abetted by statistics textbooks for biologists, there is a long history of researchers dropping an interaction effect because the interaction $p>0.05$. Don't do this. It doesn't make any sense.

1. The $p$-value is an arbitrary dichotomization of a continuous variable. Would it make sense to behave differently if the interaction were $p=0.051$ vs. $p=0.049$, given that these two p-values are effectively identical?
2. A $p$-value is not evidence that an effect is zero, or "doesn't exist", or even that an effect is "trivially small". This is because $p$-values are a function of measurement error, sampling error, and sample size, in addition to effect size.

## Example 3: Estimation of synergy ("More than the sum of the parts") (plant root growth)

This chapter uses data from an experiment measuring the effect of two defense signals on the defense response in Maize plants. In response to herbivory from insects, maize, and other plants, release multiple, chemical signals into the air (chemicals that evaporate into the air are known as volatile compounds). These chemicals signal the plant, and neighboring plants, to secrete anti-herbivory hormones, including abcisic acid and jasmonic acid. The researchers investigated the effects of two volatile compounds, (Z)‐3‐hexenyl acetate (HAC) and Indole, on the defense response both each without the other and in combination.

The example data come from Figure 1a, which is the effect of HAC and Indole on tissue concentrations of the hormone abcisic acid (ABA). The design is fully crossed with two factors, each with two levels: $hac$, with levels "HAC-" and "HAC+", and $indole$, with levels ("Indole-" and "Indole+").

```{r factorial-2way-table, echo = FALSE, message=FALSE, warning=FALSE}
df <- data.frame("HAC-" = c("Control", "Indole"),
                 "HAC+" = c("HAC", "HAC+Indole"))
row.names(df) <- c("Indole-", "Indole+")
knitr::kable(df, col.names = c("HAC-", "HAC+"))
```

```{r factorial-import-ecology, echo=FALSE, warning=FALSE, message=FALSE}

data_from <- "Integration of two herbivore-induced plant volatiles results in synergistic effects on plant defense and resistance"
file_name <- "Data for Dryad.xlsx"
file_path <- here(data_folder, data_from, file_name)

fig1 <- read_excel(file_path,
                     sheet = "Fig. 1",
                     range = "A3:K23", # 1 blank column
                     col_names = TRUE) %>%
  data.table() %>%
  clean_names()

fig1[treat == "Control", c("hac", "indole") := list("HAC-", "Indole-")]
fig1[treat == "HAC", c("hac", "indole") := list("HAC+", "Indole-")]
fig1[treat == "Indole", c("hac", "indole") := list("HAC-", "Indole+")]
fig1[treat == "HAC+Indole", c("hac", "indole") := list("HAC+", "Indole+")]

treat_levels <- c("Control", "HAC", "Indole", "HAC+Indole")
fig1[, treat := factor(treat, levels = treat_levels)]
fig1[, hac := factor(hac)] # levels in correct order
fig1[, indole := factor(indole)] # levels in correct order
# View(fig1)
```
### Analysis

#### Examine the data

```{r}
qplot(x = treat, y = aba, data = fig1)
```

Too few points for box plot. control variance is small. No obvious implausible points. fit with lm but recognize small n warning for any inference.

#### Fit the model
```{r factorial-fit-model}
m1 <- lm(aba ~ hac * indole, data = fig1)
```

#### Model check

```{r}
qqPlot(m1, id = FALSE)
```

A plausible sample from a Normal distribution.

```{r}
spreadLevelPlot(m1)
```
A bit of increase in variance with mean. 

#### Inference from the model

```{r}
m1_coef <- cbind(coef(summary(m1)),
                 confint(m1))
knitr::kable(m1_coef, digits = 3)
```


```{r}
m1_emm <- emmeans(m1, specs = c("hac", "indole"))
knitr::kable(m1_emm, digits = 2)
```

```{r}
m1_pairs <- contrast(m1_emm,
                     method = "revpairwise",
                     simple = "each",
                     combine = TRUE,
                     adjust = "none") %>%
  summary(infer = TRUE)

knitr::kable(m1_pairs, digits = 2)
```

```{r}
m1_interaction <- contrast(m1_emm,
                           interaction = c("revpairwise"),
                           by = NULL) %>%
  summary(infer = TRUE)

m1_interaction %>%
  kable(digits = c(1,1,1,2,0,2,1,2,3)) %>%
  kable_styling()
```

#### Plot the model

xxx if interested in synergy plot the coefficients! this is what we want. When would we want to plot all simple effects?

```{r factorial-aba-prepare-plot, echo=FALSE}

m1_emm_dt <- summary(m1_emm) %>%
  data.table
m1_emm_dt[, treat := levels(fig1$treat)]
m1_pairs_dt <- data.table(m1_pairs)

# create contrast column with unique value each row
within_group <- c("Indole-", "Indole+", "HAC-", "HAC+")
m1_pairs_dt[, contrast := paste(within_group, contrast, sep =": ")]

# add interaction row
m1_pairs_dt <- rbind(m1_pairs_dt,
                     data.table(contrast = "Interaction",
                                estimate = m1_coef[4, "Estimate"],
                                lower.CL = m1_coef[4, "2.5 %"],
                                upper.CL = m1_coef[4, "97.5 %"],
                                p.value = m1_coef[4, "Pr(>|t|)"]),
                     fill = TRUE)

# pvalString is from package lazyWeave
m1_pairs_dt[ , p_pretty := pvalString(p.value)]
# also create a column with "p-val: "
m1_pairs_dt[ , pval_pretty := paste("p-val:", p_pretty)]
```


```{r factorial-aba-build-plot, echo=FALSE}
# response plot using treat column
gg_response <- ggplot(data = fig1,
                            aes(x = aba,
                                y = treat,
                                color = hac,
                                shape = indole)) +
  # points
  geom_jitter(alpha = 0.5,
            height = 0.4) +
  
  # plot means and CI
  geom_errorbar(data = m1_emm_dt,
                aes(x = emmean,
                    xmin = lower.CL,
                    xmax = upper.CL,
                    color = hac),
                width = 0,
  ) +
  
  geom_point(data = m1_emm_dt,
             aes(x = emmean,
                 color = hac),
             size = 3
  ) +

  # aesthetics
  xlab("ABA (ng per g FW)") +
  scale_color_manual(values=pal_okabe_ito,
                     name = NULL) +
  theme_pubr() +
  theme(legend.position="none") +
  theme(axis.title.y=element_blank()) +
  
  NULL

# gg_effect

contrast_order <- m1_pairs_dt[, contrast]
m1_pairs_dt[, contrast := factor(contrast, contrast_order)]
contrast_labels <- c("HAC - Control",
                     "HAC+Indole - Indole",
                     "Indole - Control",
                     "HAC+Indole - HAC",
                     "Interaction")

gg_effect <- ggplot(data = m1_pairs_dt, 
                    aes(x = estimate,
                        y = contrast)) +
  # confidence level of effect
  geom_errorbar(aes(xmin=lower.CL, 
                    xmax=upper.CL),
                width=0, 
                color="black") +
  # estimate of effect
  geom_point(size = 3) +
  
  # draw a line at effect = 0
  geom_vline(xintercept=0, linetype = 2) +
  # draw a line to sep interaction from simple effects
  geom_vline(xintercept=4.5, color = "gray") +
  
  # p-value. The y coordinates are set by eye
  annotate(geom = "text",
           label = m1_pairs_dt$p_pretty,
           y = 1:5,
           x = 40,
           size = 3) +
  
  # aesthetics
  
  scale_y_discrete(labels = contrast_labels) +
  scale_x_continuous(position = "top") + # move to top

  xlab("Effect (ng per g FW)") +
  ylab("Contrast") +
  coord_cartesian(xlim = c(-10,40)) +
  
  # use ggpubr theme
  theme_pubr() +
  theme(axis.title.y=element_blank()) +
  
  NULL

#gg_effect
```

```{r factorial-aba-plot, echo=FALSE}
plot_grid(gg_effect,
          gg_response,
          nrow=2,
          align = "v",
          axis = "lr",
          rel_heights = c(0.5,1))
```
## Understanding the linear model with crossed factors 3 (The interaction effect is the measure of synergy)

### A factorial model adds an interaction effect to the coefficients table

A factorial design allows a researcher to estimate the interaction between two factors. An interaction is an effect between two variables in a linear model. The interaction effect is the coefficient $beta_3$ in the linear model

\begin{equation}
aba = \beta_0 + \beta_1 hac_{HAC^+} + \beta_2 indole_{Indole^+} + \beta_3 hac_{HAC^+}:indole_{Indole^+} +\varepsilon
(\#eq:factorial-full)
\end{equation}

$hac_{HAC^+}$ and $indole_{Indole^+}$ are dummy-coded **indicator variables** indicating group membership. $hac_{HAC^+} : indole_{Indole^+}$ is a dummy-coded variable for the interaction between $hac$ and $indole$. The value of this variable is the product of the two indicator variables in the interaction ($hac_{HAC^+}$ and $indole_{Indole^+}$), which can be verified with the model matrix (which here, is computed from the subset of the data that includeds only the first two rows of each treatment combination)

```{r factorial-dummy, echo=FALSE}
m1_mm <- model.matrix(m1)
first_two <- c(1:2, 6:7, 11:12, 16:17)
knitr::kable(m1_mm[first_two,])
```

The coefficient table is

```{r factorial-coef, echo=FALSE}
knitr::kable(m1_coef, digits=c(2,2,1,3, 2, 2), caption="Coefficient table of the factorial model")
```

1. The Intercept ($b_0$) is the mean of the reference (HAC-/Indole-) group, and so the mean of the upper left cell ("Control") in Table \@ref(tab:factorial-coef).
2. The hacHAC+ coefficient ($b_1$) is the estimate of the added HAC effect relative to the reference, and so is the mean of the lower left cell ("HAC") minus the mean of the upper left cell ("Control"). Another way of stating this is, it is **the effect of HAC when Indole is at its reference level**.
3. The indoleIndole+ coefficient ($b_2$) is the estimate of the added Indole effect relative to the reference, and so is the mean of the upper right cell ("Indole") minus the mean of the upper left cell ("Control"). Another way of stating this is, **it is the effect of Indole when HAC is at its reference level**.
4. The hacHAC+:indoleIndole+ coefficient ($b_3$) is the estimate of the **interaction effect**. If we added the HAC effect ($b_1$) and the Indole effect ($b_2$) to the Control mean ($b_0$), we would get the expected mean of the "HAC+Indole" group **if the effects were additive**. The hacHAC+:indoleIndole+ effect ($b_3$) is the additional bit that we need to get from this additive expectation to the modeled HAC+Indole mean (Figure \@ref(factorial-what-are-coefficients-plot)). It is the difference between the mean of the bottom right cell ("HAC+Indole") and the sum of the coefficients of the other three cells ($b0$, $b1$, $b2$).

An interaction is a **non-additive effect**. Think about this. Adding HAC alone increases ABA concentration by 13.3 ng per g FW. Adding Indole alone increases ABA concentration by 8.8 ng per g FW. If these effects were purely additive, then adding both HAC and Indole to the Control mean should result in a mean of 19.2 + 13.3 + 8.8 = 41.3 ng per g FW in the HAC+Indole group. The modeled mean is 54.1 ng per g FW. The difference observed - additive is 54.1 - 41.3 = 12.8 ng per g FW. Compare this to the interaction coefficient in the coefficient table.

```{r twoway-synergy-build, echo=FALSE, eval=FALSE, fig.cap="Meaning of coefficients in factorial model. b0 (blue line) is the mean of the reference (Control). b1 (orange line) is the /HAC effect. Numerically, it is the mean of the HAC group minus the mean of the reference. b2 (green line) is the Indole effect. Numerically it is the mean of the Indole group minus the mean of the reference. The expected mean of the HAC+Indole group if HAC and Indole were additive is b0 + b1 + b2 (gray circle). b3 (purple line) is the interaction effect. Numerically, it is the observed mean of the HAC+Indole group minus the expected additive mean (gray circle)"}

m1 <- lm(aba ~ hac * indole, data = fig1)
m1_emm <- emmeans(m1, specs=c("hac", "indole"))%>%
  summary() %>%
  data.table()

treatment_levels <- c("Control", "HAC", "Indole", "HAC+Indole")
m1_emm[, treatment := factor(treatment_levels, treatment_levels)]

b <- coef(m1)
b0 <- b[1]
b1 <- b[2]
b2 <- b[3]
b3 <- b[4]


gg_add <- ggplot(data = m1_emm,
                 aes(x = treatment,
                     y = emmean,
                     color = treatment)) +
  geom_point(size = 4, alpha = 0.5) +
  coord_cartesian(ylim = c(0, 60)) +

  geom_point(aes(x = 4,
                 y = b0 + b1 + b2),
             size = 4,
             alpha = 0.5) +
  
  geom_segment(aes(x = c(1, 2, 3, 4),
                   y = c(0,emmean[1],emmean[1],emmean[1]),
                   xend = c(1, 2, 3, 4),
                   yend = emmean),
               size = 1,
               show.legend = FALSE) +
 
  geom_segment(aes(x = c(4),
                   y = c(emmean[1]),
                   xend = c(4),
                   yend = c(emmean[2])),
               size = 1,
               color = pal_okabe_ito_blue[2]) +
  geom_segment(aes(x = c(4),
                   y = c(emmean[2]),
                   xend = c(4),
                   yend = c(emmean[2] + (emmean[3]-emmean[1]))),
               size = 1,
               color = pal_okabe_ito_blue[3]) +
  
  annotate(geom = "text",
           label="b[0]",
           parse=TRUE,
           x = 1.15,
           y = b0/2) +
  annotate(geom = "text",
           label="b[1]",
           parse=TRUE,
           x = 2.15,
           y = b0 + b1/2) +
  annotate(geom = "text",
           label="b[2]",
           parse=TRUE,
           x = 3.15,
           y = b0 + b2/2) +
  annotate(geom = "text",
           label="b[3]",
           parse=TRUE,
           x = 4.15,
           y = b0 + b1 + b2 + b2/2) +

  annotate(geom = "text",
           label="interaction effect",
           x = 2.55,
           y = 51.5,
           size = 4) +
  geom_segment(aes(x = 3.1,
                   y = 51,
                   xend = 3.9,
                   yend = b0+b1+b2+b3/2),
               arrow=arrow(length = unit(0.05, "npc"),
                           ends="last",type="open"),
               color="black") +

  annotate(geom = "text",
           label="additive expectation",
           x = 2.55,
           y = 44.5,
           size = 4) +
  geom_segment(aes(x = 3.2,
                   y = 44,
                   xend = 3.9,
                   yend = b0+b1+b2+0.3),
               arrow=arrow(length = unit(0.05, "npc"),
                           ends="last",type="open"),
               color="black") +
  
  scale_color_manual(values=pal_okabe_ito_blue,
                     name = "Group mean") +
  
  ylab("ABA (ng per g FW)") +
  theme_minimal_hgrid() +
  
  theme(axis.title.x=element_blank()) +
  scale_y_continuous(
    # don't expand y scale at the lower end
    expand = expansion(mult = c(0, 0.05)))  +

  NULL

image_path <- here("images", "twoway-synergy.png")
ggsave(image_path,
       width = 7,
       height = 5,
       units = "in")

gg_add
```

```{r twoway-synergy, echo=FALSE, out.width="80%", fig.cap = "Why the interaction effect is the stimulation-induced TLR9 effect"}
image_path <- here("images", "twoway-synergy.png")

knitr::include_graphics(image_path)
```

### The biological interpretation of an interaction effect

The biological reasons causing interaction effects are highly variable but lets consider how interactions *might* arise in the context of the ABA defense response to HAC and Indole. Additive effects (no interaction) may occur when combined treatments act independently of each other. This might occur in the Maize ABA response if the signaling path from HAC reception to ABA secretion and Indole reception to ABA secretion occur in different cells or by different signaling pathways and activity in either pathway has no influence on the other. Positive, or **synergistic** interaction effects may occur when combined treatments augment each other's ability to affect the response. This could occur in the Maize ABA response if an active signaling path from one of the volatile compound to ABA secretion makes the signaling path from the other compound to ABA secretion more sensitive to that compound. Negative, or **antagonistic** interaction effects may occur when combined treatments interfere with each other's ability to affect the response. This could occur in the Maize ABA response if an active signaling path from one of the volatile compound to ABA secretion inhibits the signaling path from the other compound to ABA secretion. Negative interaction effects could also occur if there is a simple threshold response and all it takes is either the HAC or the Indole signaling pathway to activate the response.


## Working in R
### Model formula

A full-factorial model with two factors is specified in the model formula as `y ~ A*B` where $texttt{A}$ is the first factor, and $texttt{B}$ is the second factor. R expands this formula to `y ~ 1 + A + B + A:B` where the colon indicates an interaction (multiplicative) effect.

```{r }
m1 <- lm(aba ~ hac * indole, data = fig1)
m1_coef <- coef(summary(m1))
  
m1_coef %>%
  kable(digits = c(1,2,2,3))
```

The additive model is specified by the formula `y ~ A + B`

```{r urchin-additive}
m2 <- lm(aba ~ hac + indole, data = fig1)
m2_coef <- coef(summary(m2))
  
m2_coef %>%
  kable(digits = c(1,2,2,5))
```

## Hidden Code
### Import exp2j (Example 1)

```{r twoway-exp2j-import-show, echo=TRUE}
data_from <- "TLR9 and beclin 1 crosstalk regulates muscle AMPK activation in exercise"
file_name <- "41586_2020_1992_MOESM4_ESM.xlsx"

file_path <- here(data_folder, data_from, file_name)

treatment_levels  <- c("WT Rest",
                       "WT Stimulation",
                       "KO Rest",
                       "KO Stimulation")
exp2j_wide <- read_excel(file_path,
                         sheet = "2j",
                         range = "A5:D13",
                         col_names = TRUE) %>%
  data.table()

colnames(exp2j_wide) <- treatment_levels

exp2j <- melt(exp2j_wide,
              measure.vars = treatment_levels,
              variable.name = "treatment",
              value.name = "glucose_uptake") %>%
  na.omit() # danger!

exp2j[, c("genotype", "stimulation") := tstrsplit(treatment,
                                                  " ",
                                                  fixed = TRUE)]

genotype_levels <- c("WT", "KO")
stimulation_levels <- c("Rest", "Stimulation")
exp2j[, genotype := factor(genotype,
                           levels = genotype_levels)]
exp2j[, stimulation := factor(stimulation,
                              levels = stimulation_levels)]
# View(exp2j)
```

