---
title: "regression to the mean"
author: "Jeffrey A. Walker"
date: "11/21/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
It is common to measure the outcome variable ($Y$) both before and after the experimental treatments are applied and then compare the pre-post *change* in $Y$ in response to the treatment using a $t$-test or ANOVA using this linear model

\begin{equation}
Y_{post}-Y_{pre} = \beta_0 + \beta_1 Treatment + \varepsilon
(\#eq:cov-change-score)
\end{equation}

**Don't do this**. Instead, add the pre-treatment measure into the model as a covariate.

\begin{equation}
Y_{post} = \beta_0 + \beta_1 Y_{pre} + \beta_2 Treatment + \varepsilon
(\#eq:ancova-4)
\end{equation}

where $Treatment$ is a dummy variable for a two-level factor. A pre-treatment measure ($Y_{pre}$) is often called the *baseline* measure. The change in $Y$ ($\Delta Y = Y{post} - Y_{pre}$) is sometimes called a change score or gain score. If you really want to estimate the treatment effect on the change from pre-treatment value to post-treatment value, then use model \@ref(eq:ancova-4) with $\Delta Y$ as the response -- the $p$-value will be precisely the same (the estimate and SE will differ of course because the response variable is different).

The reason why a researcher should not model a change score ($\Delta Y$) as a function of $Treatment$ without $Y_{pre}$ as a covariate is a phenomenon called **regression to the mean**. To explain regression to the mean, I use fake data simulated to model the results from an important study on gut microbiomes. In this study, the authors (Turnbaugh et al. xxx) showed that mice with feces from obese (genotype *ob/ob*) donors had higher weight gain than mice with feces from lean (genotype *+/+*) donors, presumably because of the differences in microbial communities between the donor types (shown elsewhere in their paper). To support the inference of a large difference in weight change, they illustrated the percent change in each treatment level in their Fig 3C, which is replicated here using simulated data generated to match the original summary statistics (Figure \@ref(fig:ancova-mouseplot1)). 

```{r ancova-mouse, echo=FALSE}
# create fake data using available summary statistics from paper.
# Use a while loop to generate a new data set each loop and 
# check how close the summary statistics of the fake data are
# to those of the Turnbaugh data. If all stats
# are within tolerance, keep and plot

# "mouse" is a table of summary statistics of the +/+
# and ob/ob groups reconstructed from data in Turnbaugh
mouse <- data.table(treatment = c('+/+', 'ob/ob'),
                    n = c(10, 9),
                    percent = c(.27, .47), # fraction not percent
                    change = c(.86, 1.3),
                    se_percent = c(0.036, 0.083),
                    se_change = c(0.1, 0.2)
                    )
mouse[, init := change/percent]
mouse[, final := init + change]
# sample sd, assume homogenous pre/post
mouse[, sd := sqrt(se_change^2*n/2)]

# reorder columns
mouse <- mouse[, .SD, .SDcols=c('treatment', 'n', 'init', 'final', 'sd', 'change', 'percent', 'se_change', 'se_percent')]

# compute some statistics for later use
init_diff <- (mouse[2, init] - mouse[1, init]) /
  sqrt(mean(mouse[, sd^2]))
final_diff <- (mouse[2, final] - mouse[1, final]) /
  sqrt(mean(mouse[, sd^2]))

# generate replica fake data

percent.lean <- mouse[treatment == '+/+', percent]
percent.obese <- mouse[treatment == 'ob/ob', percent]
percent.lean.se <- mouse[treatment == '+/+', se_percent]
percent.obese.se <- mouse[treatment == 'ob/ob', se_percent]

# parameters for model
# variances for both init and final
Sigma_ii <- mean(mouse[, sd^2])

# standard deviations
sigma <- sqrt(Sigma_ii)

# pre-post correlation. I have no idea what the actual value is.
rho <- 0.4

# initial weight
mu <- mean(mouse[, init])

# delta is the post - pre effect for the control group
delta <- mean(mouse[, change]) 

# tau is the treatment effect (on top of delta)
tau <- 0

Sigma <- matrix(c(c(Sigma_ii, rho*Sigma_ii), c(rho*Sigma_ii, Sigma_ii)), nrow=2)
N <- sum(mouse[, n])

seed <- 1673 # recovers stats
done <- FALSE
while(done==FALSE){
  set.seed(seed)
  
  # create pre-post mouse weights that are correlated
  treatment <- rep(c('+/+','ob/ob'), mouse[,n])
  weights <- rmvnorm(n=N, mean=c(mu, mu+delta), sigma=Sigma)
  weights[treatment == 'ob/ob', 2] <- 
    weights[treatment == 'ob/ob', 2] + tau
  fake_mouse <- data.table(ID = factor(1:N),
                   treatment = factor(treatment, c('+/+','ob/ob')),
                   init = weights[,1],
                   final = weights[,2])
  fake_mouse[, change := final-init]
  fake_mouse[, percent := change/init]
  fake_mouse_sum <- fake_mouse[, .(final = mean(final),
                   percent_change = mean(percent), 
                   se = sd(percent)/sqrt(.N), 
                   cs = mean(change), 
                   se.change = sd(change)/sqrt(.N)), 
               by = treatment]
  fake_mouse_sum
  if(abs(fake_mouse_sum[treatment == 'ob/ob', final] - 
         fake_mouse_sum[treatment=='+/+', final]) < 0.04 &
     abs(fake_mouse_sum[treatment == 'ob/ob', percent_change] -
         percent.obese) < 0.02 &
     abs(fake_mouse_sum[treatment == '+/+', percent_change] -
         percent.lean) < 0.02 &
     abs(fake_mouse_sum[treatment == 'ob/ob', se] -
         percent.obese.se) < 0.02 &
     abs(fake_mouse_sum[treatment == '+/+', se] -
         percent.lean.se) < 0.01 ){
    done <- TRUE
  }else{
    seed <- seed+1
  }
}

fake_mouse_sum[, percent_change := percent_change*100]
fake_mouse_sum[, se := se*100]
fake_mouse[, percent := percent*100]

```

```{r ancova-mouseplot1, echo=FALSE, fig.cap="Figure 3c of Turnbaugh *et al* 2006. This figure was generated with simulated data matching the summary statistics given in Turnbaugh *et al* 2006", fig.small = TRUE}
# plot it
gg <- ggplot(data=fake_mouse_sum, aes(x=treatment, y=percent_change, fill=treatment)) +
  geom_errorbar(aes(ymin=(percent_change-se), ymax=(percent_change+se)), width=.2) +
  geom_col(fill=c('white', 'black'), color='black') +
  ylab("Increase in Body Fat (%)") +
  xlab("Donor") +
  scale_y_continuous(limits=c(0,60)) +
  theme_minimal(base_size=18) +
  theme(legend.position='none')
gg

```

That looks like a big difference, with the mice from the obese-donor treatment level gaining much more fat than the mice from the lean-donor treatment level. Turnbaugh et al. used a simple t-test of this percent change to test the effect of the *ob/ob* treatment. The linear model underneath this $t$-test is

\begin{equation}
percent\_change\_fat = \beta_0 + \beta_1 obese + \varepsilon
\end{equation}

where $percent\_change\_fat$ is the percent change in fat from baseline and $obese$ is a dummy variable with *ob/ob* $= 1$. The percent change in fat is $\frac{fat_{post} - fat_{pre}}{fat_{pre}} \times 100$, so is a function of the change score $\Delta_{fat} = fat_{post} - fat_{pre}$.

The model coefficients are

```{r ancova-mouse-model1, echo=FALSE}
fit1 <- lm(percent ~ treatment, data=fake_mouse)
coef(summary(fit1))
confint(fit1)
```

Or, the increase in fat in the obese-treated mice was 21.9% (95%CI: 4.7, 39.2%, $p=0.016$) greater than the increase in lean-treated mice. This result, if generally verified with replication and rigorous probing, would have spectacular implications for human health.

### Regression to the mean in words

Regression to the mean is the phenomenon that if an extreme value is sampled, the next sample will likely be less extreme. This makes sense, if you randomly sample a single human male and that individual is 6'10" (about 4 standard deviations above the mean), the next human you randomly sample will almost certainly be closer to the mean human male. Or, if you randomly sample five human males and the mean height in the group is 5'1" (about 3 standard deviations below the mean), the next sample of five human males that you measure will almost certainly be closer to the mean human male.

How does regression to the mean apply to the analysis of change scores in a pre-post experiment, like the mouse fecal transplant study? In a pre-post experiment, subjects are randomized to treatment group. The response is measured at baseline and again at the conclusion of the experiment. Despite random treatment assignment, the mean fat weight of the *ob/ob* group at baseline was 1.2 standard deviations smaller than that of the *+/+* group. If there is no treatment effect, what is the expected difference at the end?

To answer this, we need to know how an individual's fat weight at the end is related to its fat weight at baseline. An individual's final fat is dependent on its initial fat if factors that contribute to the measurement of fat are the same at baseline and the end. For example, if an individual has relatively high metabolism both at baseline and at the end, then that individual might have relatively low fat at baseline and at the end. This dependence of final value on baseline value is quantified by the correlation between the two measures. This correlation is $\rho$ (the greek letter rho). Factors that change over the duration of the experiment, including random measurement error, cause the correlation to be less than one. The two extremes of this correlatioun, and the expected difference in fat weight at the end are:

1. $\rho=0$ -- if an individual's final fat is independent of its initial fat then we expect the difference at end to be zero.
2. $\rho=1$ -- if an individuals's final fat is entirely dependent on its initial fat, then we'd expect the mean fat weight of the *ob/ob* group to be 1.2 standard deviations smaller than that of the *+/+* group, exactly as it was at baseline.

Regression to the mean happens when $\rho < 1$ and its consequences increase as $\rho$ goes to zero. What is meant by "consequences"?

The fat weight of the *ob/ob* group at baseline is 1.2 standard deviations smaller than that of the *+/+* group. If $\rho=0$, then we'd expect the difference between mean fat weight at the end of the experiment to be zero. *Given the starting differences in mean weight*, to get to zero difference at the end, the *ob/ob* mice would have to gain more fat weight than the *+/+* mice. Since the expectation of the mean difference at the end is zero the expectation of the change score *must be bigger for the ob/ob mice than for the +/+ mice*. That is the expectation of the *difference* in change score is conditional on (or "a function of") the difference in fat weight at baseline.

### Regression to the mean in pictures

Let's simulate this to pump our intuition about regression to the mean and its consequences on pre-post experiments. 

1. randomly sample a normal distribution as the "initial weight" and randomly assign to treatment class
2. let the final weight have some correlation ($\rho$) with the initial weight. Some correlation should make sense -- we expect a mouse that has more fat than average at the start of the experiment to also have more fat than average at the end of the experiment. Run the experiment at different values of this correlation to see how it effects regression to the mean.
3. Do not add a treatment effect. We want to explore the behavior of the nill null hypothesis.

```{r ancova-sim1, echo=FALSE, fig.cap="Effect of initial difference in weight on the difference in change score. Increased initial difference in weight results in an increased differences in change score between treatment and control. Four different values of *rho* (the correlation between initial and final weights) were simulated. Only when *rho*=1 is there no influence of initial difference, because whatever differences occur at baseline will be perfectly preserved in the final measure. The X gives the values in the original Turnbaugh data"}
niter <- 1000
n <- 10
N <- n*2
# parameters from code block "originalPlot"
d_obs.init <- mouse[treatment=='ob/ob', init] - mouse[treatment=='+/+', init]
d_obs.change <- mouse[treatment=='ob/ob', change] - mouse[treatment=='+/+', change]
d_obs.percent <- mouse[treatment=='ob/ob', percent] - mouse[treatment=='+/+', percent]

init_diff <- numeric(niter)
cs_diff <- numeric(niter)
percent_diff <- numeric(niter)
res_mat <- data.table(NULL)
for(rho in c(0, 0.33, 0.66, 1)){
  Sigma <- matrix(c(c(sigma^2, rho*sigma^2), c(rho*sigma^2, sigma^2)), nrow=2)
  for(iter in 1:niter){
    # col 1 is initial weights, col 2 is final weights
    weights <- rmvnorm(n=N, mean=c(mu, mu+delta), sigma=Sigma)
    init <- weights[,1]
    final <- weights[,2]
    change <- final - init
    percent <- change/init
    init_diff[iter] <- mean(init[1:n]) - mean(init[(n+1):(n*2)]) 
    cs_diff[iter] <- mean(change[1:n]) - mean(change[(n+1):(n*2)]) 
    percent_diff[iter] <- mean(percent[1:n]) - mean(percent[(n+1):(n*2)]) 
  }
  res_mat <- rbind(res_mat, data.table(rho=rho, init=init_diff, cs=cs_diff, percent=percent_diff))
}

# plot it
res_mat[, rho:=factor(rho)]
labs <- levels(res_mat[, rho])
#cols <- colorblind_brewer(length(labs))
cols <- colorblind_pal
#cols <- viridis(4)[1:4]
gg <- ggplot(data=res_mat, aes(x=init, y=cs, color=rho)) +
  geom_point() +
  geom_point(aes(x=d_obs.init, y=d_obs.change), color='black', shape='X', size=5) +
  ylab("Difference in change score") +
  xlab("Initial difference") +
#  scale_color_viridis(discrete=TRUE) +
  scale_colour_manual(values=cols, labels = labs) +
  theme_minimal(base_size=14) +
  NULL
gg

# modeled p-value
p.value <- sum(abs(res_mat$cs >= d_obs.change))/(niter*4) # 

# conditional p if rho=0.33
res_mat <- rbind(res_mat, data.table(rho=0.33, init=d_obs.init, cs=d_obs.change, percent=d_obs.percent))
pfit <- lm(cs~init, data=res_mat[rho==0.33])
res <- residuals(pfit)
p.cond1 <- sum(abs(res) >= abs(res[1001]))/1001

# conditional p if rho=0.66
res_mat <- rbind(res_mat, data.table(rho=0.66, init=d_obs.init, cs=d_obs.change, percent=d_obs.percent))
pfit <- lm(cs~init, data=res_mat[rho==0.66])
res <- residuals(pfit)
p.cond2 <- sum(abs(res) >= abs(res[1001]))/1001

```

What's happening in Figure \@ref(fig:ancova-sim1)? Each point is a result for a single, simulated experiment. In total, there are 1000 simulated experiments for each of four values of $\rho$. The *x*-axis is the difference between the means of the two treatment levels at baseline (*Initial difference*). The *y*-axis is the difference in mean change score between the two treatment levels -- that is the difference in the means of $\Delta Y$ from equation \@ref(eq:ancova-5). This difference in $\Delta Y$ is the effect of the treatment the researchers are interested in. The *unconditional* expectation of this difference is zero

\begin{equation}
\mathrm{E}(\Delta Y_{ob/ob} - \Delta Y_{+/+}) = 0
\end{equation}

but the change conditional on baseline is not zero

\begin{equation}
\mathrm{E}(\Delta Y_{ob/ob} - \Delta Y_{+/+}) \ne 0
\end{equation}

Instead, the conditional expectation is a function of the difference at baseline. If the initial difference in weight happens to be unusually large and negative, the expected difference in change score is unusually positive. This non-zero expectation means that the estimate of the treatment effect is **conditionally biased** for any model that does not include the baseline fat weight as a covariate. And, from a frequentist perspective, the Type I error for a test of a difference in $\Delta Y$ is strongly dependent on the initial difference in weight.

The big X in the plot indicates the difference at baseline and difference in $\Delta Y$ for the original fecal transplant study. The difference in $Delta Y$ is unusually positive (about .6% of the $|\delta Y|$ are larger) but very close to the expected value given the unusually large, negative difference at baseline. In other words, the probability of the data, or more extreme than the data, is not 0.006 but something larger and perhaps, much larger (the computed value depends on the observed $\rho$. From, the plot, the X is very unusual if $\rho=1$, pretty unusual if $\rho=0.66$, but pretty common if $\rho=0.33$ or if $\rho=0$).

### Do not use percent change, believing that percents account for effects of initial weights

Some researchers mistakenly believe that a $t$-test of percent change automatically adjusts for effects in initial weight, since this initial weight is in the denominator of the percent. This is wrong. The dependency of the difference in change between treatments on the initial difference between treatments is more severe if change is measured as a percent, because the numerator (the change score) is expected to be larger if the denominator is smaller (initial measure). Using the simulated data from above, here is this dependency.

```{r ancova-sim2, echo=FALSE, fig.cap="Effect of initial difference in weight on the difference in percent change. Increased initial difference in weight results in an increased differences in Percent change between treatment and control. Four different values of *rho* (the correlation between initial and final weights) were simulated. Note there is no value of *rho* where the difference in percent change is independent of the initial difference. The X gives the values in the original Turnbaugh data."}
gg <- ggplot(data=res_mat, aes(x=init, y=percent, color=rho)) +
  geom_point() +
  geom_point(aes(x=d_obs.init, y=d_obs.percent), color='black', shape='X', size=5) +
  ylab("Difference in percent change") +
  xlab("Initial difference") +
  scale_colour_manual(values=cols, labels = labs) +
  theme_minimal(base_size=14) +
  NULL
gg

```

