---
title: "simulation -- area under the curve"
author: "Jeffrey A. Walker"
date: "1/31/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(here)

library(nlme)
library(lmerTest)
library(emmeans)
library(MASS)
library(mvtnorm)
library(Hmisc) # auc using trap.rule, probably faster than my code

library(knitr)
library(kableExtra)

here <- here::here


```

Rho is the conditional error correlation matrix.
```{r simulation functions}
  rho.1.2 <- 0.5
  rho.1.p <- 0.6
  rho.max.max <- 0.8
  rho.max.min <- 0.7
  
fake_Rho <- function(p=5,
                     rho.base.2 = 0.6, # cor(t_0, t_1)
                     rho.base.p = 0.5, # cor(t_0, t_p)
                     rho.max.max = 0.8, # cor(t_p-1, t_p)
                     rho.max.min = 0.7, # cor(t_1, t_2)
                     rho.min = 0.5){
  # rho.base.2 and rho.base.p control the correlations of baseline with post-baseline measures. These tend to be lower than the correlations among post-baseline measures, generally between 0 and 0.5
  # rho.max.max and rho.max.min control the maximum post-baseline correlations. In general the correlations are highest beteween succesive times and are highest between final two times, which is rho.max.max. rho.max.min is the correlation between time 2 and 3. The correlations drop to rho.min.
  
  Rho_fake <- matrix(1, nrow=p, ncol=p)
  for(i in 1:(p-1)){
    cells <- p - i
    row.max <- (cells-1)/(p-2-1)*rho.max.min +
      (1-(cells-1)/(p-2-1))*rho.max.max
    inc <- -(row.max - rho.min)/(p - 2 -1)
    for(j in (i+1):p){
      if(i==1){
        Rho_fake[i,j] <- (j-2)/(p-2)*rho.base.p +
          (1-(j-2)/(p-2))*rho.base.2
        Rho_fake[j,i] <- Rho_fake[i,j]
      }else{
        Rho_fake[i,j] <- row.max + inc*(cells - (p-j+1))
        Rho_fake[j,i] <- Rho_fake[i,j]
      }
    }
  }
  return(Rho_fake)
}
```


```{r}
# x - the time points. Weights constructed from these
# y - the values that we want to compute the mean
weighted_mean <- function(x, y){
  p <- length(x)
  pm1 <- p-1

  x_0 <- c(0, (x[2:p] - x[1:pm1]))
  x_1 <- c((x[2:p] - x[1:pm1]), 0)
  weights <- (x_0 + x_1)/2
  
  w_mean <- (weights %*% y)/sum(weights)
  return(w_mean)
}

```

```{r}
# about the variables
  n_sim = 4000 # number of iterations in simulation
  n_levels = c(10, 10) # sample size fore each group
  treatment_levels = c("cn", "tr") # names of groups - don't change
  R_max = c(0.6, 0.8) # baseline max, non-baseline max for the matrix of correlations between time points. See the information on constructing Rho
  times = c(0, 15, 30, 60, 120) # measurement time periods (minutes)
  mu = c(150, 150, 150, 150, 150) # glucose for cn at baseline carried over for each time. These should all be the same
  beta_time = c(0, 80, 80, 50, 0) # effect of glucose at each post-baseline time. Baseline included in vector - should be zero. add to mu. This is what creates the curve for cn
  beta_trt = c(20, 20, 20, 20, 20) # effect of treatment at baseline (carried over to all times). Add this to mu + beta_time. This creates the curve for tr if there is no differential response to glucose.
  beta_ixn = c(0, 20, 20, 20, 10) # differential effect of glucose for tr. Add this to mu + beta_time + beta_trt. This creates the curve for tr. The first value should be zero because ixn doesn't add to this.
  sigma = c(10, 25, 25, 25, 25) # sd of sample at each time
  seed_i = 1 # seed

simulator <- function(
  n_sim = 4000,
  n_levels = c(10, 10),
  treatment_levels = c("cn", "tr"),
  R_max = c(0.6, 0.8), 
  times = c(0, 15, 30, 60, 120),
  mu = c(150, 150, 150, 150, 150),
  beta_time = c(0, 80, 80, 50, 0),
  beta_trt = c(20, 20, 20, 20, 20),
  beta_ixn = c(0, 20, 20, 20, 10),
  sigma = c(10, 25, 25, 25, 25),
  seed_i = 1
){
  set.seed(seed_i)
  p <- length(mu)
  N <- sum(n_levels)
  baseline_max <- R_max[1]
  non_baseline_max <- R_max[2]
  R <- fake_Rho(p,
                rho.base.2 = baseline_max,
                rho.base.p = 3/4*baseline_max,
                rho.max.max = non_baseline_max,
                rho.max.min = 7/8*non_baseline_max,
                rho.min = 5/8*non_baseline_max)
  
  R_0 <- R[,1] # correlation of each time with time0
  R_0[1] <- 0 # don't add this component to first beta
  Sigma <- diag(sigma)%*%R%*%diag(sigma)
  
  # period durations
  n_times <- length(times)
  full_period <- times[n_times] - times[1]
  post_period <- times[n_times] - times[2]

    # init fake data matrix
  fake_data <- data.table(
    treatment = factor(rep(c("cn", "tr"), each = n_levels))
    )
  fake_data[, id:=factor(1:.N)]
  
  param_list <- c("r_obs",
                  "b_tr_0",
                  "b_change",
                  "b_ancova",
                  "b_ancova_post_full",
                  "b_ancova_post")
  method_list <- c("change", "change_post", "ancova", "ancova_post_full", "ancova_post")
  res <- matrix(as.numeric(NA),
                 nrow = n_sim,
                 ncol = length(c(param_list, method_list))) %>%
    data.table()
  colnames(res) <- c(param_list, method_list)

  for(iter in 1:n_sim){
    Y <- rbind(rmvnorm(n_levels[1],
                       mu + beta_time,
                       Sigma),
               rmvnorm(n_levels[2],
                       mu + beta_time + beta_trt + beta_ixn,
                       Sigma))
    Y1 <- matrix(c(Y[, 1],
                   rep(0, N*(n_times-1))),
                 ncol = n_times)
    fake_data[, glucose_0 := Y[,1]]
    fake_data[, auc := apply(Y, 1, trap.rule,
                             x = times)]  
    fake_data[, auc_post_full := apply(Y-Y1, 1, trap.rule,
                                  x = times)]
    fake_data[, auc_post := apply(Y[, -1], 1, trap.rule,
                                  x = times[-1])]
    fake_data[, auc_change := apply(Y-Y[,1], 1, trap.rule,
                                    x = times)]
    fake_data[, auc_change_post := apply((Y-Y[,1])[,-1], 1, trap.rule,
                                         x = times[-1])]
    
    # convert to means over the period
    fake_data[, glucose_mean := auc/full_period]
    fake_data[, glucose_mean_post_full := auc_post_full/full_period]
    fake_data[, glucose_mean_post := auc_post/post_period]
    fake_data[, glucose_change := auc_change/full_period]
    fake_data[, glucose_change_post := auc_change_post/post_period]
    
    
    # compute correlated response
    fake_data[, glucose_0_res := glucose_0 - mean(glucose_0), by = treatment]
    fake_data[, auc_res := auc - mean(auc), by = treatment]
    res[iter, r_obs := cor(fake_data[, glucose_0_res], fake_data[, auc_res])]
    
    # change score full period
    m1 <- lm(glucose_change ~ treatment,
             data = fake_data)
    # change score post period
    m2 <- lm(glucose_change_post ~ treatment,
             data = fake_data)
    # ANCOVA LM full period
    m3 <- lm(glucose_mean ~ glucose_0 + treatment,
             data = fake_data)
    # ANCOVA LM post period
    m4 <- lm(glucose_mean_post_full ~ glucose_0 + treatment,
             data = fake_data)
     # ANCOVA LM post period
    m5 <- lm(glucose_mean_post ~ glucose_0 + treatment,
             data = fake_data)
   
    m1_coef <- coef(summary(m1))
    m2_coef <- coef(summary(m2))
    m3_coef <- coef(summary(m3))
    m4_coef <- coef(summary(m4))
    m5_coef <- coef(summary(m5))
    
    res[iter, b_tr_0 := mean(fake_data[treatment == "tr", glucose_0]) -
      mean(fake_data[treatment == "cn", glucose_0])]
    res[iter, b_change := m1_coef["treatmenttr", "Estimate"]]
    res[iter, b_ancova := m3_coef["treatmenttr", "Estimate"]]
    res[iter, b_ancova_post_full := m4_coef["treatmenttr", "Estimate"]]
    res[iter, b_ancova_post := m5_coef["treatmenttr", "Estimate"]]
    
    res[iter, change := m1_coef["treatmenttr", "Pr(>|t|)"]]
    res[iter, change_post := m2_coef["treatmenttr", "Pr(>|t|)"]]
    res[iter, ancova := m3_coef["treatmenttr", "Pr(>|t|)"]]
    res[iter, ancova_post_full := m4_coef["treatmenttr", "Pr(>|t|)"]]
    res[iter, ancova_post := m5_coef["treatmenttr", "Pr(>|t|)"]]
  }
  
  return(res)

}


```


# Global variables

```{r}
do_it <- TRUE
sim_i <- 0 # this will increment up
n_sim_g = 4000
n_levels_g = c(10, 10)
treatment_levels_g = c("cn", "tr")
R_max_g = c(0.6, 0.8)
times_g = c(0, 15, 30, 60, 120)
mu_g = c(150, 150, 150, 150, 150)
beta_time_g = c(0, 80, 80, 50, 0)
beta_trt_i = c(20, 20, 20, 20, 20) # local
beta_ixn_i = c(0, 20, 20, 20, 10) # local
sigma_g = c(10, 25, 25, 25, 25)
seed_g = 1

p_values <- data.table(NULL)
method_list <- c("change",
                 "change_post",
                 "ancova",
                 "ancova_post_full",
                 "ancova_post")

```

# randomized at baseline -- Type I

```{r}
sim_i <- sim_i + 1
baseline <- "randomized"
test <- "type I"
beta_trt_i = c(0, 0, 0, 0, 0)
beta_ixn_i = c(0, 0, 0, 0, 0)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_g,
    n_levels = n_levels_g,
    treatment_levels = treatment_levels_g,
    R_max = R_max_g,
    times = times_g,
    mu = mu_g,
    beta_time = beta_time_g,
    beta_trt = beta_trt_i,
    beta_ixn = beta_ixn_i,
    sigma = sigma_g,
    seed_i = seed_g
  )
  p_values <- rbind(p_values,
                    data.table(
                      sim = sim_i,
                      baseline = baseline,
                      test = test,
                      sim_res
                    ))
  apply(sim_res[, .SD, .SDcols = method_list],
        2,
        function(x) sum(x < 0.05)/n_sim_g)
}

```

# randomized at baseline -- Power

```{r}
sim_i <- sim_i + 1
baseline <- "randomized"
test <- "power"
beta_trt_i = c(0, 0, 0, 0, 0)
beta_ixn_i = c(0, 20, 20, 20, 10)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_g,
    n_levels = n_levels_g,
    treatment_levels = treatment_levels_g,
    R_max = R_max_g,
    times = times_g,
    mu = mu_g,
    beta_time = beta_time_g,
    beta_trt = beta_trt_i,
    beta_ixn = beta_ixn_i,
    sigma = sigma_g,
    seed_i = 2
  )
  p_values <- rbind(p_values,
                    data.table(
                      sim = sim_i,
                      baseline = baseline,
                      test = test,
                      sim_res
                    ))
  apply(sim_res[, .SD, .SDcols = method_list],
        2,
        function(x) sum(x < 0.05)/n_sim_g)
  trap.rule(times,beta_ixn_i)/(times[5] - times[1]) # treatment effect
  trap.rule(times,beta_ixn_i)/(times[5] - times[2]) # treatment effect
  trap.rule(times[-1],beta_ixn_i[-1])/(times[5] - times[2]) # treatment effect
  mean(sim_res$b_change)
  mean(sim_res$b_ancova)
  mean(sim_res$b_ancova_post_full)
  mean(sim_res$b_ancova_post)
  qplot(x = b_tr_0,
        y = b_ancova,
        data = sim_res)
  qplot(x = b_tr_0,
        y = b_ancova_post,
        data = sim_res)
  coef(summary(lm(b_ancova ~ b_tr_0,
        data = sim_res)))
  coef(summary(lm(b_ancova_post ~ b_tr_0,
        data = sim_res)))

}

```

# Different groups at baseline -- Type I

```{r}
sim_i <- sim_i + 1
baseline <- "different"
test <- "type I"
beta_trt_i = c(20, 20, 20, 20, 20)
beta_ixn_i = c(0, 0, 0, 0, 0)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_g,
    n_levels = n_levels_g,
    treatment_levels = treatment_levels_g,
    R_max = R_max_g,
    times = times_g,
    mu = mu_g,
    beta_time = beta_time_g,
    beta_trt = beta_trt_i,
    beta_ixn = beta_ixn_i,
    sigma = sigma_g,
    seed_i = seed_g
  )
  p_values <- rbind(p_values,
                    data.table(
                      sim = sim_i,
                      baseline = baseline,
                      test = test,
                      sim_res
                    ))
  apply(sim_res[, .SD, .SDcols = method_list],
        2,
        function(x) sum(x < 0.05)/n_sim_g)
}

```

# Different groups at baseline -- Power

```{r}
sim_i <- sim_i + 1
baseline <- "different"
test <- "power"
beta_trt_i = c(20, 20, 20, 20, 20)
beta_ixn_i = c(0, 20, 20, 20, 10)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_g,
    n_levels = n_levels_g,
    treatment_levels = treatment_levels_g,
    R_max = R_max_g,
    times = times_g,
    mu = mu_g,
    beta_time = beta_time_g,
    beta_trt = beta_trt_i,
    beta_ixn = beta_ixn_i,
    sigma = sigma_g,
    seed_i = seed_g
  )
  p_values <- rbind(p_values,
                    data.table(
                      sim = sim_i,
                      baseline = baseline,
                      test = test,
                      sim_res
                    ))
  apply(sim_res[, .SD, .SDcols = method_list],
        2,
        function(x) sum(x < 0.05)/n_sim_g)

  trap.rule(times,beta_ixn_i)/(times[5] - times[1]) # treatment effect
  trap.rule(times,beta_ixn_i)/(times[5] - times[2]) # treatment effect
  trap.rule(times[-1],beta_ixn_i[-1])/(times[5] - times[2]) # treatment effect
  mean(sim_res$b_change)
  mean(sim_res$b_ancova)
  mean(sim_res$b_ancova_post_full)
  mean(sim_res$b_ancova_post)
}

```
# read/write

```{r}
sim_data_folder <- "sim_data"
sim_fn <- "auc.Rds"
sim_path <- here(sim_data_folder, sim_fn)
if(do_it == TRUE){
  saveRDS(p_values, sim_path)
}else{
  p_values <- readRDS(sim_path)
}

```

# table
```{r}
p_values[, lapply(.SD,
                  function(x) sum(x < 0.05)/mean(n_sim_g)),
         .SDcols = method_list,
         by = c("sim", "baseline","test")] %>%
  kable(digits = c(1,1,1,3,3,3,3)) %>%
  kable_styling()

```
# correlated response
```{r}
p_values[, .(lower = quantile(r_obs, 0.1),
             median = quantile(r_obs, 0.5),
             upper = quantile(r_obs, 0.9)),
         by = .(baseline, test)]
# same each sim because restarted seed each sim so using same random error
```
# sim 1 results

Conditional bias in the estimation of the treatment effect (b_ixn) when the treatment is randomized at baseline, and the consequence of this on p-values. The consequence on p-values for this parameterization (E(r) = 0.67) is way too trivial to care for empirical reasons.

```{r}
qplot(x = b_tr_0, y = b_tr_change, data = p_values[sim == 1])
qplot(x = abs(b_tr_0), y = change, data = p_values[sim == 1]) +
  geom_smooth()

fit <- lm(change ~ r_obs + abs(b_tr_0), data = p_values[sim == 1])
coef(summary(fit))
```

# sim 2 results

```{r}
beta_ixn_i <- c(0, 20, 20, 20, 10) # for sim 2
trap.rule(times,beta_ixn_i)/(times[5] - times[1])
mean(p_values[sim == 2, b_tr_change])
mean(p_values[sim == 2, b_tr_ancova])
qplot(x = b_tr_0, y = b_tr_change, data = p_values[sim == 2]) +
  geom_smooth()
qplot(x = abs(b_tr_0), y = change, data = p_values[sim == 2]) +
  geom_smooth()


```
# sim 3 results

```{r}
qplot(x = b_tr_0, y = b_tr_change, data = p_values[sim == 3])
qplot(x = b_tr_0, y = b_tr_change, data = p_values[sim == 3])

qplot(x = abs(b_tr_0), y = change, data = p_values[sim == 3]) +
  geom_smooth()

fit <- lm(change ~ r_obs + abs(b_tr_0), data = p_values[sim == 3])
coef(summary(fit))
fit <- lm(change_post ~ r_obs + abs(b_tr_0), data = p_values[sim == 3])
coef(summary(fit))

```

# sim 4 results

```{r}
beta_ixn_i <- c(0, 20, 20, 20, 10) # for sim 4
trap.rule(times,beta_ixn_i)/(times[5] - times[1])
mean(p_values[sim == 4, b_tr_change])
mean(p_values[sim == 4, b_tr_ancova])
qplot(x = b_tr_0, y = b_tr_change, data = p_values[sim == 4]) +
  geom_smooth()
qplot(x = b_tr_0, y = b_tr_ancova, data = p_values[sim == 4]) +
  geom_smooth()
qplot(x = abs(b_tr_0), y = change, data = p_values[sim == 4]) +
  geom_smooth()
qplot(x = abs(b_tr_0), y = ancova, data = p_values[sim == 4]) +
  geom_smooth()

```

