# Simulating linear models
## Setup

```{r simulation-ttest-setup, include=FALSE, echo=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(data.table)
library(mvtnorm)
library(magrittr)

library(ggplot2)
library(ggpubr)
library(knitr)
library(kableExtra)
library(printy)

sim_data <- "sim_data"

```

## Simulating a mean and variance in a single group
### Using the error draw specification as a generating model

$$
\begin{equation}
y_i = \beta_0 + \varepsilon_i\\
\varepsilon \sim \mathrm{Normal}(0, \sigma^2)
\end{equation}
$$



```{r}
#1. create simulation and model parameters and give these parameters values
n = 10 # sample size
beta_0 = 22 # mean
sigma = 1.5 # std deviation

#2. Add the generating code
e <- rnorm(n, mean = 0, sd = sigma) # error draw
fake_y <- beta_0 + e # generating function

#3 compare the estimates to the parameters
mean(fake_y)
sd(fake_y)
```

### Using the conditional draw specification as a generating model

$$
\begin{equation}
y_i \sim \mathrm{Normal}(\mu_i, \sigma^2)\\
\mu_i = \beta_0\\
\end{equation}
$$

```{r}
#1. create simulation and model parameters and give these parameters values
n = 10 # sample size
beta_0 = 22 # mean
sigma = 1.5 # std deviation

#2. Add the generating code
mu <- beta_0
fake_y <- rnorm(n, mean = mu, sd = sigma) # conditional draw

#3 compare the estimates to the parameters
mean(fake_y)
sd(fake_y)
```
Notes

1. The advantage of the conditional draw is that this is how fake data are generated for non normal data. For example, if we want to generate fake count data we do this

```{r}
mu <- beta_0
fake_count <- rpois(n, lambda = mu) # conditional draw
```

and not this

```{r, eval=FALSE}
fake_count_wrong <- beta_0 + rpois(n, lambda = ?)

```

2. That said, I almost always use the error draw when I simulate from the normal distribution because inertia.

## Embed the fake data generating code within a function

```{r error_draw}
error_draw <- function(
  n = 10, # sample size
  beta_0 = 10, # mean
  sigma = 1 # std deviation
){
  e <- rnorm(n, mean = 0, sd = sigma)
  y <- beta_0 + e
  return(y)
}
```


```{r}
conditional_draw <- function(
  n = 10, # sample size
  beta_0 = 10, # mean
  sigma = 1 # std deviation
){
  mu <- beta_0
  fake_y <- rnorm(n, mean = mu, sd = sigma) # conditional draw
  return(y)
}
```

Notes

1. The functions have default values for the arguments n, beta_0, and sigma. 

## Use the function

```{r}
set.seed(1) # comment out if you want a different sample

# parameterize
n <- 11
beta_0 <- 22
sigma <- 2.5

# get the fake data
fake_y <- error_draw(
  n = n,
  beta_0 = beta_0,
  sigma = sigma
)

# compare the estimates to the parameters
mean(fake_y)
sd(fake_y)
```

## Use the function a bunch of times
### Analyses within loop

```{r}

n_sim <- 1000

# parameterize
n <- 11
beta_0 <- 22
sigma <- 2.5

# set memory for values we are keeping

stats_list <- c("mean_i", "sd_i")
stats_dt <- matrix(as.numeric(NA),
                       nrow = n_sim,
                       ncol = length(stats_list)) %>%
  data.table
setnames(stats_dt, old = names(stats_dt), new = stats_list)

for(sim_i in 1:n_sim){
  set.seed(sim_i) # comment out if you want a different sample
  fake_y <- error_draw(
    n = n,
    beta_0 = beta_0,
    sigma = sigma
  )
  stats_dt[sim_i, mean_i := mean(fake_y)]
  stats_dt[sim_i, sd_i := sd(fake_y)]
}

# compare the estimates to the parameters
stats_dt[, .(mean = mean(mean_i),
             sd = sd(sd_i))]

```

## Simulate a simple linear model with a single factor with two levels ("t-test")

```{r}

```

