# Simulations -- Count data (alternatives to a t-test) {#count-sim}

```{r setup}
library(here)
library(data.table)
library(readxl)

library(MASS)
library(lmPerm)
library(nlme)
library(glmmTMB)
library(emmeans)

library(knitr)
library(kableExtra)

here <- here::here
```

## Use data similar to Figure 6f from Example 1

The simulated data are modeled to look like the [Figure 6f](#violations-6f) data used in Example 1 of the Violations chapter.

```{r sim-counts-fig6f-import}
data_folder <- "data"
data_from <- "Exercise reduces inflammatory cell production and cardiovascular inflammation via instruction of hematopoietic progenitor cells"
file_name <- "41591_2019_633_MOESM8_ESM.xlsx"
file_path <- here(data_folder, data_from, file_name)

# assuming mice are independent and not same mouse used for all three treatment
melt_col_names <- c("Sedentary", "Exercise")
fig6f <- read_excel(file_path,
                     sheet = "Figure 6f",
                     range = "A7:B29",
                     col_names = TRUE) %>%
  data.table() %>%
  melt(measure.vars = melt_col_names,
       variable.name = "treatment",
       value.name = "neutrophils") %>%
  na.omit() # danger!

treatment_levels <- melt_col_names
fig6f[, treatment := factor(treatment,
                            levels = treatment_levels)]

# neutrophils is count/10^6
fig6f[, neutrophil_count := round(neutrophils*10^6, 0)]

# fig6f[1, neutrophil_count]
#View(fig6f)
```

```{r sim-counts-fig6f-,1}
m1 <- glmmTMB(neutrophil_count ~ treatment,
              data = fig6f,
              family = nbinom2(link = "log"))
mu_cn_obs <- exp(coef(summary(m1))$cond["(Intercept)", "Estimate"])
theta_obs <- sigma(m1) %>%
  round(1)

# compare with glm.nb, which is used for the simulation
m2 <- glm.nb(neutrophil_count ~ treatment,
              data = fig6f)
mu_cn_obs
m2$theta
```

## Functions

```{r simulator}
simulator <- function(n_sim = 1000,
                      n_levels = c(10, 10),
                      treatment_levels = c("cn", "tr"),
                      mu_levels = c(10, 10),
                      theta_levels = c(1, 1),
                      sigma_levels = c(1,1),
                      normal = FALSE,
                      seed = 1){
  
  set.seed(seed)
  
  N <- sum(n_levels)
  
  mu_sim <- rep(mu_levels, n_levels)
  theta_sim <- rep(theta_levels, n_levels)
  sigma_sim <- rep(sigma_levels, n_levels)
  
  method_list <- c("lm", "gls", "glm_nb", "glm_qp", "lmp", "mww")
  p_values <- matrix(as.numeric(NA),
                     nrow = n_sim,
                     ncol = length(method_list)) %>%
    data.table()
  colnames(p_values) <- method_list
  
  fake_counts <- matrix(as.numeric(NA),
                        nrow = N,
                        ncol = n_sim)
  
  if(normal == TRUE){
    for(j in 1:n_sim){
      fake_counts[, j] <- rnorm(N,
                                mean = mu_sim,
                                sd = sigma_sim) %>%
        round(0)
    }
  }else{
    for(j in 1:n_sim){
      fake_counts[, j] <- rnegbin(N,
                                  mu = mu_sim,
                                  theta = theta_sim)
    }
  }
  
  for(j in 1:n_sim){
    fake_data[, count := fake_counts[,j]]
    m1 <- lm(count ~ treatment, data = fake_data)
    m2 <- gls(count ~ treatment,
              data = fake_data,
              weights = varIdent(form = ~ 1 | treatment))
    m2_pairs <- contrast(emmeans(m2, specs = "treatment"),
                         method = "revpairwise") %>%
      summary()
    m3 <- glm.nb(count ~ treatment, data = fake_data)
    m4 <- glm(count ~ treatment,
                 data = fake_data,
                 family = quasipoisson)
    # m3 <- glmmTMB(count ~ treatment,
    #                data = fake_data,
    #                family = nbinom2(link = "log"))$cond
    # m4 <- glmmTMB(count ~ treatment,
    #                data = fake_data,
    #                family = nbinom1(link = "log"))$cond
    m5 <- lmp(count ~ treatment,
              data = fake_data,
              perm = "Exact",
              settings = FALSE)
    m6 <- wilcox.test(count ~ treatment,
                      data = fake_data) 
    p_values[j, lm := coef(summary(m1))[2, "Pr(>|t|)"]]
    p_values[j, gls := m2_pairs[1, "p.value"]]
    p_values[j, glm_nb := coef(summary(m3))[2, "Pr(>|z|)"]]
    p_values[j, glm_qp := coef(summary(m4))[2, "Pr(>|t|)"]]
    p_values[j, lmp := coef(summary(m5))[2, "Pr(Prob)"]]
    p_values[j, mww := m6$p.value]
  }
  
  return(p_values)
}
```

```{r}
binder <- function(
  p_values,
  sim_res,
  sim_i,
  n_levels_sim,
  mu_levels_sim,
  theta_levels_sim,
  sigma_levels_sim,
  normal_sim){
  
  sim_table <- data.table(
    sim = sim_i,
    n_sim = n_sim_i,
    n1 = n_levels_sim[1],
    n2 = n_levels_sim[2],
    n = ifelse(n_levels_sim[1] == n_levels_sim[2],
               "=",
               "!="),
    mu1 = mu_levels_sim[1],
    mu2 = mu_levels_sim[2],
    type = ifelse(mu_levels_sim[1] == mu_levels_sim[2],
                  "type 1",
                  "power"),
    theta1 = theta_levels_sim[1],
    theta2 = theta_levels_sim[2],
    sigma1 = sigma_levels_sim[1],
    sigma1 = sigma_levels_sim[2],
    normal = normal_sim,
    sim_res    
  )
  
  p_values <- rbind(p_values, sim_table)
  return(p_values)
}

```

## Simulations

```{r sim-counts-initialize}
do_it <- FALSE
p_values <- data.table(NULL)
sim_i <- 0
n_sim_i = 4000


```

### Type I, Pseudo-Normal distribution

```{r, warning=FALSE, message=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(10, 10)
mu_levels_sim <- c(1000, 1000)
theta_levels_sim <- c(NA, NA) # not used
sigma_levels_sim <- c(100, 100)
normal_sim <- TRUE
if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Type I, neg binom, equal n

```{r, message=FALSE, warning=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(10, 10)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs)
theta_levels_sim <- c(theta_obs, theta_obs)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Type I, neg binom, equal n, small theta

```{r, message=FALSE, warning=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(10, 10)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs)
theta_levels_sim <- c(1, 1)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Type I, neg binom, unequal n

```{r, message=FALSE, warning=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(12, 8)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs)
theta_levels_sim <- c(theta_obs, theta_obs)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Power, Pseudo-Normal distribution, equal n

```{r, warning=FALSE, message=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(10, 10)
mu_levels_sim <- c(1000, 1100)
theta_levels_sim <- c(NA, NA) # not used
sigma_levels_sim <- c(100, 100)
normal_sim <- TRUE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
  
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Power, neg binom, equal n

```{r, message=FALSE, warning=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(10, 10)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs*.7)
theta_levels_sim <- c(theta_obs, theta_obs)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Power, neg binom, small theta

```{r, message=FALSE, warning=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(10, 10)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs*.7)
theta_levels_sim <- c(1, 1)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Power, neg binom, unequal n

```{r, message=FALSE, warning=FALSE}

sim_i <- sim_i + 1
n_levels_sim <- c(12, 8)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs*.7)
theta_levels_sim <- c(theta_obs, theta_obs)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Power, neg binom, unequal n, unequal theta

What if the treatment affects the variance and the mean?

```{r, message=FALSE, warning=FALSE}

sim_i <- sim_i + 1
n_levels_sim <- c(12, 8)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs*.7)
theta_levels_sim <- c(1, theta_obs)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

### Type 1, neg binom, equal n, unequal theta

What if the treatment affects the variance but not the mean?

```{r, message=FALSE, warning=FALSE}
sim_i <- sim_i + 1
n_levels_sim <- c(12, 8)
mu_levels_sim <- c(mu_cn_obs, mu_cn_obs)
theta_levels_sim <- c(theta_obs, 1)
sigma_levels_sim <- c(NA, NA) # not used
normal_sim <- FALSE

# fake_data has to be global for emmeans to run inside 
# a function
treatment_levels <- c("cn", "tr")
fake_data <- data.table(
  treatment = factor(rep(treatment_levels, n_levels_sim),
                     levels = treatment_levels)
)

if(do_it == TRUE){
  sim_res <- simulator(
    n_sim = n_sim_i,
    n_levels = n_levels_sim,
    mu_levels = mu_levels_sim,
    theta_levels = theta_levels_sim, # not used
    sigma_levels = sigma_levels_sim,
    normal = normal_sim
  )
  p_values <- binder(p_values,
                     sim_res,
                     sim_i,
                     n_levels_sim,
                     mu_levels_sim,
                     theta_levels_sim,
                     sigma_levels_sim,
                     normal_sim)
#  apply(sim_res, 2, function(x) sum(x < 0.05)/n_sim_i)
}

```

## Save it, Read it

```{r}
sim_data_folder <- "sim_data"
sim_fn <- "counts_t_test_alternative.Rds"
sim_path <- here(sim_data_folder, sim_fn)
if(do_it == TRUE){
  saveRDS(p_values, sim_path)
}else{
  p_values <- readRDS(sim_path)
}

```

## Analysis

```{r}
method_list <- c("lm", "gls", "glm_nb", "glm_qp", "lmp", "mww")

p_values[, lapply(.SD,
                  function(x) sum(x < 0.05)/mean(n_sim)),
         .SDcols = method_list,
         by = c("sim", "normal","type", "n", "theta1", "theta2")] %>%
  kable(digits = c(0,0,0,0,2,2,3,3,3,3,3,3)) %>%
  kable_styling()

```

